---
title: "Columns & headers"
description: >
  The column surface: name and configure every column with cols() and
  col_spec(), then build multi-level header bands with headers().
vignette: >
  %\VignetteIndexEntry{Columns & headers}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

Two verbs shape the columns of a table. `cols()` names each data column
and hands it a `col_spec()`; `headers()` groups those columns under
multi-level spanning bands. This article covers both.

## Naming columns with `cols()`

`cols()` takes one named argument per data-frame column. The name is the
column in your data; the value is a `col_spec()` describing how to show
it. Columns you do not name keep their defaults.

```{r}
#| label: basic-cols
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

::: {.callout-warning}
## Column order is the data frame's, not the call's

Columns render in the order they appear in the **data frame**, not the
order you list them in `cols()`. To reorder, reorder the data first:

```r
demo <- saf_demo[c("variable", "stat_label", "placebo", "drug_50", "drug_100", "Total")]
```
:::

## The `col_spec()` toolbox

Every column option lives on `col_spec()`:

| Argument | What it does |
|---|---|
| `label` | Column-header text (use `\n` for multi-line headers) |
| `align` | `"left"`, `"center"`, `"right"`, or `"decimal"` |
| `usage` | `"display"` (default), `"group"`, `"indent"`, or `"id"` |
| `group_display` | For a group column: `"header_row"`, `"column"`, `"column_repeat"` |
| `visible` | `FALSE` keeps the column in the data but off the page |
| `na_text` | String shown for `NA` cells (default `""`) |
| `width` | `"auto"`, a number (inches), or a unit string like `"3cm"` |
| `indent_by` | Name of a column holding per-row indent depth |
| `format` | A formatter applied to the column's values |

### Decimal alignment

`align = "decimal"` lines numbers up on the decimal point using real font
metrics, so a column of `53 (62.1%)` and `8 (9.3%)` stays aligned in
print. It is the workhorse alignment for clinical value columns.

### Group columns: section headers or a repeated column

A `usage = "group"` column does not become an ordinary column. By
default (`group_display = "header_row"`) its values become **bold
section-header rows**:

```{r}
#| label: group-header-row
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),                # bold section rows
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

Switch to `group_display = "column"` to keep the group label as a real
left-hand column instead of section rows — useful when each group has a
single row.

## Hiding the columns that drive the display

Many clinical tables carry columns that control sorting, indentation, or
styling but should not appear on the page. Set `visible = FALSE`: the
column stays in `spec@data` (so `sort_rows()`, `indent_by`, and `style()`
predicates can still use it) but is not rendered.

The AE-by-SOC/PT table is the canonical case. Its `row_type`,
`indent_level`, and sort-key columns are hidden, and the preferred terms
are indented under their system-organ-class header via `indent_by`:

```{r}
#| label: ae-hidden
# first two system-organ-class blocks, to keep the preview short
ae <- saf_aesocpt[1:14, ]

tabular(
  ae,
  titles = "Table 14.3.1  Adverse Events by System Organ Class and Preferred Term"
) |>
  cols(
    soc          = col_spec(visible = FALSE),
    row_type     = col_spec(visible = FALSE),
    indent_level = col_spec(visible = FALSE),
    n_total      = col_spec(visible = FALSE),
    soc_n        = col_spec(visible = FALSE),
    label        = col_spec(label = "System Organ Class / Preferred Term",
                            indent_by = "indent_level"),
    placebo      = col_spec(label = "Placebo",  align = "decimal"),
    drug_50      = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100     = col_spec(label = "Drug 100", align = "decimal"),
    Total        = col_spec(label = "Total",    align = "decimal")
  ) |>
  style(bold = TRUE, .at = cells_body(where = row_type %in% c("overall", "soc")))
```

## BigN in the header

Treatment-group denominators belong in the column headers. Pull them
from the bundled `saf_n` and build the labels with `sprintf()`:

```{r}
#| label: bign
n <- stats::setNames(saf_n$n, saf_n$arm_short)

tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = sprintf("Placebo\n(N=%d)",  n["placebo"]),  align = "decimal"),
    drug_50    = col_spec(label = sprintf("Drug 50\n(N=%d)",  n["drug_50"]),  align = "decimal"),
    drug_100   = col_spec(label = sprintf("Drug 100\n(N=%d)", n["drug_100"]), align = "decimal"),
    Total      = col_spec(label = sprintf("Total\n(N=%d)",    n["Total"]),    align = "decimal")
  )
```

::: {.callout-tip}
## Why a data frame for BigN, not an attribute?

`saf_n` is a discoverable data frame, not an attribute on `saf_demo`,
because attributes get silently stripped by `dplyr` and base-R
subsetting. Keying the labels off `saf_n$arm_short` keeps the
denominators robust to upstream data wrangling.
:::

## Multi-level headers with `headers()`

`headers()` builds spanning bands above the leaf columns. A named
character vector spans those columns under one label:

```{r}
#| label: header-single
n <- stats::setNames(saf_n$n, saf_n$arm_short)

base <- tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = sprintf("Placebo\n(N=%d)",  n["placebo"]),  align = "decimal"),
    drug_50    = col_spec(label = sprintf("Drug 50\n(N=%d)",  n["drug_50"]),  align = "decimal"),
    drug_100   = col_spec(label = sprintf("Drug 100\n(N=%d)", n["drug_100"]), align = "decimal"),
    Total      = col_spec(label = sprintf("Total\n(N=%d)",    n["Total"]),    align = "decimal")
  )

base |>
  headers(
    "Treatment Group" = c("placebo", "drug_50", "drug_100", "Total")
  )
```

Nest a `list()` to put a spanner over a subset, leaving other columns as
**passthrough leaves** (bare strings, no band above them):

```{r}
#| label: header-nested
base |>
  headers(
    "Treatment Group" = list(
      "placebo",                                    # passthrough leaf
      "Active Drug" = c("drug_50", "drug_100"),     # sub-spanner
      "Total"                                       # passthrough leaf
    )
  )
```

::: {.callout-note}
## Spanned columns must be contiguous

A spanner sits above adjacent columns, so the columns named in one band
must be next to each other in the data frame. If `drug_50` and
`drug_100` are separated by another column, reorder the data first.
:::

## Going deeper

- **Widths.** `width = "auto"` lets the engine size the column to its
  content. Pass a number for inches (`width = 1.8`) or a unit string
  (`width = "3cm"`) to fix it.
- **Formatters.** `format` accepts a function applied to the column's
  values before rendering — handy when the upstream data is numeric and
  you want display rounding at render time.
- **Where next.** [Rows, grouping & pagination](rows-grouping-pagination.html)
  covers row order and section groups; [Styling](styling.html) targets
  any header band or column with `style()` + `cells_headers()`.
