---
title: "Rows, grouping & pagination"
description: >
  Control the order of display rows, partition a table into labelled
  subgroups, and govern how it splits across printed pages.
vignette: >
  %\VignetteIndexEntry{Rows, grouping & pagination}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

`cols()` shapes the columns; three more verbs shape the rows.
`sort_rows()` orders them, `subgroup()` partitions the table into labelled
blocks, and `paginate()` decides how it breaks across pages.

## Ordering rows with `sort_rows()`

`sort_rows()` sets the display order. Sort by any column — including
hidden ones — and the engine applies it during Resolve.

```{r}
#| label: sort-desc
# Display cells are formatted text ("217 (85.4)"), which would sort
# lexically. Derive a hidden numeric key upstream and sort on that.
ae <- saf_aeoverall
ae$total_n <- as.integer(sub(" .*", "", ae$Total))

tabular(
  ae,
  titles = "Adverse-event categories, ordered by total frequency"
) |>
  cols(
    stat_label = col_spec(label = "Category"),
    total_n    = col_spec(visible = FALSE),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  ) |>
  sort_rows(by = "total_n", descending = TRUE)
```

::: {.callout-tip}
## Per-key direction and hidden sort keys

`descending` can be a vector matching `by`, so you can sort one key down
and another up — `sort_rows(by = c("soc_n", "label"), descending = c(TRUE, FALSE))`
gives the canonical "system organ classes by descending frequency, then
preferred terms alphabetically". Sort keys are frequently `visible = FALSE`
columns that exist only to drive the order.
:::

Sorting is factor-aware: a factor column sorts by its level order, not
alphabetically, and `NA`s sort last. That is how a CDISC best-overall-
response column comes out in CR, PR, SD, PD order rather than alphabetical.

## Section groups, revisited

A `usage = "group"` column turns into bold section-header rows
(`group_display = "header_row"`). That is the row-grouping you already
saw in [Columns & headers](columns-and-headers.html): each distinct group
value introduces its block of rows.

```{r}
#| label: group-rows
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

## Partitioning with `subgroup()`

Where a group column stacks blocks *within* one table, `subgroup()`
splits the table into separate, banner-labelled partitions — each begins
on its own page in the print backends. Pass the partitioning column(s)
and a glue-style `label` template referencing those columns:

```{r}
#| label: subgroup
tabular(
  saf_subgroup,
  titles = "Table 14.2.3  Vital Signs by Sex and Age Group"
) |>
  cols(
    sex        = col_spec(visible = FALSE),
    agegr      = col_spec(visible = FALSE),
    sex_n      = col_spec(visible = FALSE),
    agegr_n    = col_spec(visible = FALSE),
    paramcd    = col_spec(visible = FALSE),
    param      = col_spec(usage = "group", group_display = "column",
                          label = "Parameter"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  ) |>
  subgroup(by = c("sex", "agegr"), label = "{sex}, {agegr}")
```

Each `sex × agegr` combination becomes its own partition with a centred
banner like *F, <65*. The columns named in `by` and in the `label`
template are hidden automatically.

::: {.callout-note}
## Group vs subgroup

A **group** (`usage = "group"`) stacks blocks within one continuous
table. A **subgroup** (`subgroup()`) cuts the table into independent
partitions that each start a new page. Use a group for "Age, then Sex,
then Race"; use a subgroup for "the whole table, repeated per sex".
:::

## Pagination with `paginate()`

`paginate()` governs how the resolved table breaks across printed pages.
You do not set rows-per-page directly — the engine computes the row
budget from the active preset (paper size, orientation, margins, font
size) and the table's chrome (titles, headers, footnotes).

```{r}
#| label: paginate
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  ) |>
  paginate(
    keep_together  = "variable",                          # never split a characteristic
    repeat_content = c("titles", "headers", "footnotes"), # repeat chrome each page
    orphan_floor   = 3,
    widow_floor    = 2
  )
```

What the arguments do:

- **`keep_together`** — names `usage = "group"` columns whose blocks must
  not be split across a page break (keep a system-organ class with its
  preferred terms).
- **`repeat_content`** — which chrome repeats at the top/bottom of every
  page; titles, headers, and footnotes by default.
- **`orphan_floor` / `widow_floor`** — minimum rows left at the bottom of
  a page / carried to the next, to avoid stranded single rows.
- **`panels`** — split a too-wide table into horizontal panels (groups of
  columns) stacked down the page.
- **`continuation`** — an optional "(continued)" marker on carried-over
  blocks.

::: {.callout-warning}
## Pagination is a print concept

The HTML backend is a single continuous document, so `paginate()` has no
visible effect in the live previews above — it shapes the **RTF, PDF, and
DOCX** output, where pages are real. Build and preview in HTML, then emit
to a paginated backend to see the splits.
:::

## Where to next

- **[Styling](styling.html)** — target rows, groups, and subgroup banners
  with `style()`.
- **[Clinical cookbook](https://vthanik.github.io/tabular/articles/clinical-cookbook.html)** — full AE-by-SOC/PT and
  subgrouped vitals tables, sorted and paginated end to end.
