---
title: "Styling"
description: >
  One style() verb plus a cells_*() location drives every visual rule;
  presets carry the cosmetic defaults, and style templates package a
  reusable house style.
vignette: >
  %\VignetteIndexEntry{Styling}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)

base <- tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

Styling in `tabular` is one verb: `style()`. You pass it visual
attributes (bold, colour, borders, padding, …) and a **location** that
says *where* to apply them. Locations are the `cells_*()` helpers.

## The shape: attributes + a location

Section-header rows are already bold by default, so to *see* a layer land
we tint their background instead:

```{r}
#| label: first-style
base |>
  style(background = "#eef2ff", .at = cells_group_headers())
```

The `.at` argument names the surface. Everything else is the styling to
apply there. Stack `style()` calls to layer rules; later layers win per
attribute.

## Where: the location helpers

| Location | Targets |
|---|---|
| `cells_body()` | Data cells (filter with `i`, `j`, or `where`) |
| `cells_headers()` | Column-header band rows (filter with `level`) |
| `cells_group_headers()` | Bold section-header rows |
| `cells_subgroup_labels()` | The subgroup banner rows |
| `cells_title()` | The title block |
| `cells_footnotes()` | The footnote block |
| `cells_pagehead()` / `cells_pagefoot()` | Page header / footer slots |
| `cells_table()` | Table-wide borders and frame (`side =`) |

```{r}
#| label: locations
base |>
  style(background = "#eef2ff", .at = cells_group_headers()) |>
  style(italic = TRUE, .at = cells_body(j = "Total")) |>
  style(border_bottom = brdr("thick"), .at = cells_headers())
```

## What: the styling attributes

A `style()` call accepts any of these:

- **Text** — `bold`, `italic`, `underline`, `color`, `background`,
  `font_family`, `font_size`.
- **Alignment** — `halign` (`"left"`/`"center"`/`"right"`), `valign`.
- **Borders** — `border_top`, `border_bottom`, `border_left`,
  `border_right`, each taking a `brdr()`.
- **Spacing** — `padding_top`/`_bottom`/`_left`/`_right`,
  `blank_above`, `blank_below`.
- **Surrounding text** — `pretext`, `posttext`.

### Borders with `brdr()`

`brdr()` describes one line. Its first argument is the **width**
(`"hairline"`, `"thin"`, `"medium"`, `"thick"`, or a number in points),
then the line `style`, then the `color`:

```{r}
#| label: borders
base |>
  style(border_top = brdr("thick"), border_bottom = brdr("thick"),
        .at = cells_headers()) |>
  style(border_bottom = brdr(width = "medium", style = "dashed"),
        .at = cells_group_headers())
```

## Data-driven styling

`cells_body(where = ...)` evaluates a predicate against the data, so you
can highlight cells by content. Reference any column — including hidden
ones:

```{r}
#| label: predicate
base |>
  style(background = "#eef6ff", .at = cells_body(where = stat_label == "n"))
```

Use `i` and `j` for positional targeting instead of a predicate:
`cells_body(j = "Total")` styles the Total column, `cells_body(i = 1:3)`
the first three rows.

## Presets: cosmetic defaults for the whole table

Where `style()` targets specific cells, `preset()` carries table-wide
geometry and cosmetic defaults — paper size, orientation, font, and five
named-list "knobs" for alignment, rules, fonts, colours, and padding:

```{r}
#| label: preset
base |>
  preset(
    font_size   = 9,
    orientation = "landscape",
    rules  = list(midrule = brdr("thick")),
    colors = list(header = c(text = "#1d4ed8", background = "#f3f4f6"))
  )
```

::: {.callout-note}
## The `c(...)` knob shape

Each cosmetic knob is a **named list keyed by surface**, and each
per-surface spec is a flat named vector — `fonts = list(body = c(family
= "Times New Roman", size = 9))`, `colors = list(header = c(text =
"#1f2937", background = "#f3f4f6"))`, `padding = list(body = c(top = 2,
bottom = 2))`. The knobs are strictly validated: unknown surfaces,
unknown keys, or nested lists are rejected with a clear error.
:::

### Session defaults with `set_preset()`

`preset()` applies to one spec in the pipe. `set_preset()` sets a
**session default** that every subsequent `tabular()` inherits — the
`ggplot2::theme_set()` analogue. `get_preset()` returns it; calling
`set_preset()` with no arguments clears it.

```{r}
#| label: set-preset
set_preset(font_size = 8)        # every later table starts at 8pt
get_preset()
set_preset()                      # clear it again
```

### A minimal theme

`preset_minimal()` strips the look down to a single column-header rule
with normal-weight text throughout — the spare house style some groups
prefer:

```{r}
#| label: minimal
base |>
  preset_minimal()
```

## Packaging a house style with `style_template()`

A submission renders dozens of tables that must share one identity. Build
the rules once with `style_template()` (chaining the *same* `style()`
verb), then attach it to a preset so every table inherits it:

```{r}
#| label: template
# Headers and section rows are bold by default, so a house style earns
# its keep with VISIBLE identity: a shaded column-header band, tinted
# section rows, and thick rules above and below the header.
house <- style_template() |>
  style(background = "#f3f4f6", .at = cells_headers()) |>
  style(background = "#eef2ff", .at = cells_group_headers()) |>
  style(border_top = brdr("thick"), border_bottom = brdr("thick"),
        .at = cells_headers())

is_style_template(house)

base |>
  preset(.style = house)
```

Attach it once per session with `set_preset(.style = house)` and every
subsequent table picks up the house style with no per-table `style()`.

## Inline markup with `md()` and `html()`

Wrap label, title, footnote, or cell text in `md()` for Markdown (bold,
italic, super/subscript) or `html()` for a constrained HTML subset:

```{r}
#| label: inline
tabular(
  saf_demo,
  titles = c(
    "Table 14.1.1",
    md("Demographic Characteristics^a^")
  ),
  footnotes = md("^a^ Body mass index summarised in kg/m^2^.")
) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

Here the `^a^` marker and its note are hand-typed and kept in sync by
you. When you want the marker assigned and placed *automatically* —
deduped across cells, byte-identical across backends — use [`footnote()`]
instead (see the *Auto-numbered footnotes* recipe in the
[clinical cookbook](articles/clinical-cookbook.html)).

## Verbatim whitespace

Significant ASCII spaces in labels and cells are preserved by default, so a
hand-built indent renders exactly as typed across every backend (HTML, RTF,
LaTeX, PDF, DOCX, Markdown). A single interior space stays breakable, so cells
still wrap; leading, trailing, and interior runs of two or more spaces become
non-breaking. Decimal padding is never affected.

```{r}
#| label: whitespace
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "     Placebo\n(N=86)", align = "decimal"),
    drug_50    = col_spec(label = "     Drug 50\n(N=96)", align = "decimal"),
    drug_100   = col_spec(label = "    Drug 100\n(N=72)", align = "decimal"),
    Total      = col_spec(label = "       Total\n(N=254)", align = "decimal")
  )
```

Set `preset(whitespace = "collapse")` to opt out and let each backend fold
space runs natively.

::: {.callout-tip}
## The cascade

Styling resolves low-to-high priority: backend defaults → session preset
(`set_preset()`) → spec preset (`preset()`) → per-spec `style()` layers.
Later layers override earlier ones per attribute, and `NA` fields leave
the prior value in place. So a house style sets the baseline and a
per-table `style()` overrides just what it names.
:::

## Where to next

- **[Clinical cookbook](clinical-cookbook.html)** — these tools applied
  to six complete production tables.
- **[Fonts & fidelity](fonts-and-fidelity.html)** — fonts, decimal
  alignment, and why the print backends are the source of truth.
