---
title: "Get started with tabular"
description: >
  Your first submission-grade clinical table in ten minutes: from a
  pre-summarised data frame to a live preview and five output formats.
vignette: >
  %\VignetteIndexEntry{Get started with tabular}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`tabular` turns a **pre-summarised data frame** into a regulatory
**table, listing, or figure** (a "TFL") and renders it natively to RTF,
PDF, HTML, LaTeX, and DOCX. This article gets you from raw numbers to a
finished table in about ten minutes.

::: {.callout-note}
## New to clinical tables?

A *TFL* is the formatted output that goes into a clinical study report or
a regulatory submission — for example, a demographics summary or an
adverse-event table. The numbers are computed upstream (by `cards`,
`gtsummary`, `dplyr`, or SAS); `tabular`'s job is to *present* them to
the exacting layout standards reviewers expect.
:::

## The one rule: bring summarised data

`tabular` does **no** statistics. You hand it a wide data frame where
**one row is one display row** and the columns are already the values you
want to show. The bundled `saf_demo` dataset is exactly this shape:

```{r}
#| label: setup
library(tabular)

head(saf_demo, 8)
```

Each row is a statistic for one characteristic (`variable`), with one
column per treatment arm. That is all `tabular` needs.

## Your first table

Wrap the data in `tabular()`, then describe each column with `cols()` and
`col_spec()`. End the chain with the spec itself to see a live preview:

```{r}
#| label: first-table
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo"),
    drug_50    = col_spec(label = "Drug 50"),
    drug_100   = col_spec(label = "Drug 100"),
    Total      = col_spec(label = "Total")
  )
```

Two things already happened for free: `usage = "group"` turned the
`variable` column into **bold section headers** (Age, Sex, Race, …)
instead of a repeated column, and the table rendered as live HTML right
here in the page.

::: {.callout-tip}
## What is that object?

Every verb returns a `tabular_spec` — an immutable description of the
table, not the table itself. Nothing is rendered until you print it (as
above) or call `emit()`. You can inspect a spec at any point in the pipe
with `print()`; piping further never mutates the previous spec.
:::

## Align the numbers

Clinical tables align numbers on the decimal point. Set
`align = "decimal"` on the value columns and `tabular` pads with real
font metrics so the columns stay aligned in print, not just on screen:

```{r}
#| label: aligned
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

## Add the submission chrome

Real tables carry a title block, BigN denominators in the column
headers, a spanning header over the treatment arms, and footnotes. Add
titles and footnotes in `tabular()`, a spanner with `headers()`, and pull
the per-arm denominators from the bundled `saf_n`:

```{r}
#| label: full-table
n <- stats::setNames(saf_n$n, saf_n$arm_short)

tab <- tabular(
  saf_demo,
  titles = c(
    "Table 14.1.1",
    "Demographic and Baseline Characteristics",
    "Safety Population"
  ),
  footnotes = c(
    "Percentages are based on the number of subjects per treatment group.",
    "BMI = body mass index."
  )
) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = sprintf("Placebo\n(N=%d)",  n["placebo"]),  align = "decimal"),
    drug_50    = col_spec(label = sprintf("Drug 50\n(N=%d)",  n["drug_50"]),  align = "decimal"),
    drug_100   = col_spec(label = sprintf("Drug 100\n(N=%d)", n["drug_100"]), align = "decimal"),
    Total      = col_spec(label = sprintf("Total\n(N=%d)",    n["Total"]),    align = "decimal")
  ) |>
  headers(
    "Treatment Group" = c("placebo", "drug_50", "drug_100", "Total")
  )

tab
```

That is a submission-shaped demographics table, built from one pipe.

## Render to five formats

The same `tab` emits to any backend. `emit()` dispatches on the file
extension (or pass `format =` explicitly):

```{r}
#| label: emit
out <- file.path(tempdir(), "t_14_1_1")

emit(tab, paste0(out, ".rtf"))    # RTF 1.9.1
emit(tab, paste0(out, ".docx"))   # native OOXML
emit(tab, paste0(out, ".html"))   # self-contained HTML
emit(tab, paste0(out, ".tex"))    # tabularray LaTeX
```

```{r}
#| label: emit-check
#| echo: false
files <- list.files(tempdir(), pattern = "^t_14_1_1", full.names = TRUE)
data.frame(
  file  = basename(files),
  bytes = file.size(files)
)
```

::: {.callout-note}
## The preview is not the deliverable

The live HTML above is a faithful *preview*. The artefact you ship is the
paginated RTF, PDF, or DOCX — same spec, same numbers, but laid out for
the page with repeated headers, footnotes, and the four-section
submission layout. See
[Fonts & fidelity](https://vthanik.github.io/tabular/articles/fonts-and-fidelity.html) for why the print backends
are the source of truth.
:::

## Where to next

- **[Core concepts](core-concepts.html)** — the mental model the rest of
  the guide assumes (the wide-data contract, the page anatomy, the
  three-phase pipeline).
- **[Columns & headers](columns-and-headers.html)** — every `col_spec()`
  option and multi-level header bands.
- **[Rows, grouping & pagination](rows-grouping-pagination.html)** — sort
  order, section groups, subgroups, and page splits.
- **[Styling](styling.html)** — `style()`, presets, and house styles.
- **[Clinical cookbook](https://vthanik.github.io/tabular/articles/clinical-cookbook.html)** — six complete
  production tables end to end.
