---
title: "Get started with tabular"
vignette: >
  %\VignetteIndexEntry{Get started with tabular}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

`tabular` turns one **pre-summarised, wide data frame** into a publication-ready
table and renders it natively to **RTF, HTML, DOCX, PDF/LaTeX and Markdown** —
from a single spec, with no Java or Office dependency.

Two things to internalise up front:

1. **tabular is display-only.** It never aggregates, filters, or computes
   statistics. You bring a summarised data frame (one input row = one display
   row); tabular lays it out and renders it. (Producing that frame from a cards
   ARD is the [*Data in*](data-in.html) article.)
2. **One immutable spec, built with verbs.** You pipe a `tabular()` object
   through verbs (`cols()`, `headers()`, `paginate()`, `preset()`, …) and finish
   with `emit()`. Each verb returns a new spec; nothing renders until `emit()`.

## Your first table

Start from a wide frame — here the bundled demographics summary, one row per
statistic, one column per treatment arm:

```{r}
data(cdisc_saf_demo, package = "tabular")
head(cdisc_saf_demo)
```

Describe the columns and emit:

```{r}
spec <- tabular(
  cdisc_saf_demo,
  titles = c("Table 14-2.01",
             "Demographic and Baseline Characteristics", "ITT Population")
) |>
  cols(
    variable   = col_spec(usage = "group", group_display = "header_row", label = ""),
    stat_label = col_spec(usage = "indent", label = "")
  )

out <- tempfile(fileext = ".html")
emit(spec, out)         # dispatches on the file extension
file.exists(out)
```

That is the whole loop: **wide frame → `tabular()` → `cols()` → `emit()`**.
Swap the extension to `.rtf`, `.docx`, `.pdf`, or `.md` and the same spec renders
to that backend.

## The pipeline at a glance

```
            cards / cardx                     tabular
  data ───► aggregate ──► pivot_across() ──► tabular() |> cols() |> … |> emit() ──► file
            (ARD, long)    (long → wide)        (one spec, any backend)
```

## Where to next

The rest of the docs are task-oriented — read the one that matches what you are
doing:

- **[Data in](data-in.html)** — turn a cards/cardx ARD (or any long ARD) into
  the wide frame, with `pivot_across()`.
- **[Structure](structure.html)** — columns, headers, BigN, and splitting wide
  or long tables across pages (`cols()`, `headers()`, `paginate()`,
  `subgroup()`).
- **[Presentation](presentation.html)** — titles, footnotes, running headers,
  and cell styling (`footnote()`, `preset()`, `style()`).
- **[Output & qualification](output.html)** — the backends, their system
  requirements, and the CDISC-pilot cross-backend validation.
