---
title: "Architecture"
description: >
  How tabular works under the hood: the immutable spec, the six-phase
  resolve engine, backend dispatch, and the QC and audit artefacts.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

`tabular` is a three-phase pipeline. Understanding the phases explains why
the API is shaped the way it is and where to look when something does not
render as expected.

![Build the spec, resolve it to a grid, emit to any backend.](../man/figures/workflow.svg){fig-alt="Pre-summarised data flows through Build, Resolve, and Emit to five output formats"}

## Phase 1 — Build

The verbs (`tabular()`, `cols()`, `headers()`, `sort_rows()`,
`subgroup()`, `paginate()`, `style()`, `preset()`) accumulate an
**immutable** `tabular_spec`, an S7 object. Each verb validates its
arguments, constructs one new property value, and returns a new spec with
that property updated. Nothing is rendered.

```{r}
#| label: build
sp <- tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(align = "decimal"),
    drug_50    = col_spec(align = "decimal"),
    drug_100   = col_spec(align = "decimal"),
    Total      = col_spec(align = "decimal")
  )

is_tabular_spec(sp)
```

Because specs are immutable, the same base can branch into many variants
without interference, and a spec can be inspected at any point in the
pipe with `print()`.

## Phase 2 — Resolve

The engine turns a spec into a `tabular_grid` by running six phases in a
fixed order:

1. **sort** — apply the `sort_rows()` order (factor-aware, NA-last).
2. **headers** — validate and flatten the header tree into bands.
3. **style** — resolve `style()` layers and predicates to per-cell nodes.
4. **format** — apply per-column formatters, NA substitution, and parse
   `md()` / `html()` inline markup.
5. **decimal** — pad `align = "decimal"` columns using real font metrics.
6. **paginate** — split the rows (and panels) into pages per the active
   preset's geometry.

The result is backend-agnostic: only the next phase differs per format.
Inspect the resolved grid without writing a file using `as_grid()`:

```{r}
#| label: resolve
g <- as_grid(sp)
is_tabular_grid(g)
```

::: {.callout-note}
## One Resolve, five Emits

The HTML you preview and the RTF you ship come from the *same* resolved
grid. The numbers, structure, alignment, and pagination are decided once
in Resolve; a backend only serialises them. That is why preview and
deliverable cannot disagree.
:::

## Phase 3 — Emit

`emit()` runs Resolve, then dispatches to a backend by the file's
extension (or an explicit `format =`):

| Extension | Backend | Notes |
|---|---|---|
| `.md`, `.markdown` | GFM pipe table | quick look; renders on GitHub |
| `.html`, `.htm` | self-contained Bootstrap HTML | the live preview |
| `.rtf` | RTF 1.9.1 (native) | submission deliverable |
| `.tex`, `.latex` | `tabularray` LaTeX | fragment or standalone |
| `.pdf` | PDF via `tinytex` | paginated print |
| `.docx` | native OOXML | no Word automation |

```{r}
#| label: emit
#| eval: false
emit(sp, "table.rtf")                 # by extension
emit(sp, "table.out", format = "docx") # or force the format
```

## QC and audit artefacts

`emit()` can drop two by-products alongside the render, for traceability:

- **`data_file`** — writes the *resolved* wide data frame (post-engine)
  next to the render, so QC can diff the numbers that were actually
  displayed. Pass a path or a function that returns one.
- **`manifest = TRUE`** — writes a CDISC ARS audit manifest describing the
  display, using the LinkML property names from the ARS logical data
  model.

```{r}
#| label: qc
#| eval: false
emit(
  sp,
  "t_14_1_1.rtf",
  data_file = "t_14_1_1_qc.csv",   # resolved data for QC
  manifest  = TRUE                  # t_14_1_1.audit.yml
)
```

::: {.callout-tip}
## The package boundary

`tabular` owns *display* only — resolve and emit. It never filters,
aggregates, weights, or inserts rows. That boundary is deliberate: it
keeps the spec a faithful description of what you handed in, which is
exactly what the QC `data_file` and the ARS manifest document.
:::

## Where to next

- **[Fonts & fidelity](fonts-and-fidelity.html)** — the metrics the
  decimal phase relies on.
