---
title: "A submission table, end to end"
description: >
  One recipe from a pre-summarised frame to a shippable RTF: nested
  section groups, per-page BigN under each subgroup banner, a running
  page header, group-aware pagination, and a QC sidecar.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

The other articles take one verb at a time. This one builds a single
deliverable — *Vital Signs by Sex*, the kind of safety table a programmer
assembles from a shell on a Monday morning — and walks it from the
pre-summarised frame all the way to the RTF you would hand to a reviewer.
Each step adds one verb; the last few add the submission chrome that turns
a table into a page.

## 1. The pre-summarised frame

`tabular` does no statistics. The summary is already computed: one row per
display row, one column per arm. `saf_subgroup` is a vital-signs summary,
split by sex and age group, with hidden columns that drive the display.

```{r}
#| label: peek
str(saf_subgroup, vec.len = 2)
```

`sex` and `agegr` are the partitioning factors; `paramcd` orders the
parameters; `param` and `stat_label` are the visible row labels; the four
arm columns hold the formatted cells. The `*_n` columns carry
denominators we will pull in later.

## 2. Columns and nested sections

`cols()` names every column. Hide the machinery (`sex`, `agegr`, the `*_n`
denominators, the `paramcd` sort key); make `agegr` and `param` section
groups so the body reads *age group → parameter → statistics*; align the
arm columns on the decimal point.

```{r}
#| label: cols
vitals <- tabular(
  saf_subgroup,
  titles = c(
    "Table 14.2.1",
    "Vital Signs Summary by Sex",
    "Safety Population"
  ),
  footnotes = "Mean (SD) unless otherwise stated. Based on observed cases."
) |>
  cols(
    sex        = col_spec(visible = FALSE),
    sex_n      = col_spec(visible = FALSE),
    agegr_n    = col_spec(visible = FALSE),
    paramcd    = col_spec(visible = FALSE),
    agegr      = col_spec(usage = "group", label = "Age Group"),
    param      = col_spec(usage = "group", label = "Parameter"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )

vitals
```

The two `usage = "group"` columns nest: each age group introduces its
block, and each parameter introduces its statistics within that block.

## 3. Split by sex with `subgroup()`

The table should repeat *per sex*, each partition starting its own page in
the print backends. `subgroup()` cuts it, and a glue-style `label`
template names each banner:

```{r}
#| label: subgroup
vitals <- vitals |>
  subgroup(by = "sex", label = "Sex: {sex}")

vitals
```

`F` and `M` now head their own partitions. The `sex` column named in `by`
is hidden automatically.

## 4. Per-page BigN

Each sex partition needs its *own* denominators — the *(N=)* on the column
headers must count the subjects in *that* page, not the whole study. Hand
`subgroup()` a `big_n` table: one row per subgroup level, one column per
arm. (These come from your ADSL counts upstream; here they sum to the
overall `saf_n` totals.)

```{r}
#| label: per-page-bign
sex_n <- data.frame(
  sex      = factor(c("F", "M"), levels = c("F", "M")),
  placebo  = c(50L, 36L),
  drug_50  = c(52L, 44L),
  drug_100 = c(38L, 34L),
  Total    = c(140L, 114L)
)

vitals <- vitals |>
  subgroup(
    by        = "sex",
    label     = "Sex: {sex}",
    big_n     = sex_n,
    big_n_fmt = "\n(N={n})"
  )

vitals
```

Now the *Sex: F* page carries *Placebo (N=50)* while the *Sex: M* page
carries *Placebo (N=36)*. In the paper backends (RTF / PDF / DOCX) the
`(N=)` rides the repeating column header on every page of the partition;
in this continuous HTML preview it shows as an *(N=)* row under each
banner.

## 5. A running header and footer

Submission pages carry chrome outside the table — the protocol and page
number at the top, the program path and run time at the foot. Set them on
the preset; `{page}` / `{npages}` resolve per page, and `{program_path}` /
`{datetime}` resolve at render time.

```{r}
#| label: chrome
vitals <- vitals |>
  preset(
    pagehead = list(
      left  = "Protocol: ABC-12345",
      right = "Page {page} of {npages}"
    ),
    pagefoot = list(
      left  = "{program_path}",
      right = "{datetime}"
    )
  )
```

## 6. Group-aware pagination

`paginate()` governs the page breaks. You do not set rows-per-page — the
engine sizes the budget from the preset and chrome. Keep each parameter's
statistics together, and repeat the titles, headers, and footnotes on
every page.

```{r}
#| label: paginate
vitals <- vitals |>
  paginate(
    keep_together  = "param",
    repeat_content = c("titles", "headers", "footnotes")
  )
```

## 7. Emit, with a QC sidecar

The spec is complete. `emit()` dispatches on the file extension. Pass
`data_file` to drop the resolved wide frame beside the render — the QC
programmer reads it back and checks the numbers independently.

```{r}
#| label: emit
out <- tempfile(fileext = ".rtf")
qc  <- tempfile(fileext = ".csv")

emit(vitals, out, data_file = qc)

file.exists(out)   # the submission deliverable
file.exists(qc)    # the post-engine wide frame, for double programming
```

The **same** `vitals` spec emits to every backend — swap the extension and
the per-page BigN, the running header, and the pagination all carry across
unchanged:

```{r}
#| label: backends
for (ext in c(".rtf", ".html", ".tex", ".docx")) {
  cat(ext, "->", basename(emit(vitals, tempfile(fileext = ext))), "\n")
}
```

`.pdf` works the same way; it compiles through `xelatex`, so it needs a
TeX install (`check_latex()` reports readiness — see the
[README](https://vthanik.github.io/tabular/#installation)).

## Where this fits

This is the whole pipeline in one place: `tabular()` to start,
`cols()` to shape, `subgroup()` to partition and carry per-page BigN,
`preset()` for the running chrome, `paginate()` for the breaks, and
`emit()` to ship — with `data_file` leaving a QC trail. From here:

- [Columns & headers](columns-and-headers.html) — the full `col_spec()`
  toolbox and multi-level header bands.
- [Rows, grouping & pagination](rows-grouping-pagination.html) — section
  groups, the running-header API, and pagination in depth.
- [Clinical cookbook](clinical-cookbook.html) — more complete production
  tables, sorted and paginated end to end.
