---
title: "Clinical cookbook"
description: >
  Six complete, submission-shaped tables built end to end: demographics,
  adverse events, vital signs, best overall response, and a subgrouped
  analysis. Copy a recipe and swap in your own summarised data.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)

saf <- stats::setNames(saf_n$n, saf_n$arm_short)   # safety BigN
eff <- stats::setNames(eff_n$n, eff_n$arm_short)    # efficacy BigN
```

Each recipe is a full pipeline on bundled demo data. The table you see is
the live HTML preview; in practice you would end the chain with
`emit(tab, "t_x.rtf")` to produce the submission deliverable. Swap the
demo dataset for your own pre-summarised wide frame and the recipe holds.

## 1. Demographics and baseline characteristics

The canonical safety-population summary: a group column for the
characteristic, decimal-aligned arm columns with BigN, a treatment-group
spanner, and a denominator footnote.

```{r}
#| label: demographics
demo <- saf_demo[c("variable", "stat_label", "placebo", "drug_50", "drug_100", "Total")]

tabular(
  demo,
  titles = c(
    "Table 14.1.1",
    "Demographic and Baseline Characteristics",
    sprintf("Safety Population (N=%d)", saf["Total"])
  ),
  footnotes = c(
    "Percentages are based on the number of subjects per treatment group.",
    "BMI = body mass index (kg/m^2)."
  )
) |>
  cols(
    variable   = col_spec(usage = "group", label = "Characteristic"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = sprintf("Placebo\n(N=%d)",  saf["placebo"]),  align = "decimal"),
    drug_50    = col_spec(label = sprintf("Drug 50\n(N=%d)",  saf["drug_50"]),  align = "decimal"),
    drug_100   = col_spec(label = sprintf("Drug 100\n(N=%d)", saf["drug_100"]), align = "decimal"),
    Total      = col_spec(label = sprintf("Total\n(N=%d)",    saf["Total"]),    align = "decimal")
  ) |>
  headers("Treatment Group" = c("placebo", "drug_50", "drug_100", "Total"))
```

## 2. Overall adverse-event summary

A compact high-level AE table, ordered by descending total frequency so
the most common categories lead.

```{r}
#| label: ae-overall
ae0 <- saf_aeoverall[c("stat_label", "placebo", "drug_50", "drug_100", "Total")]

tabular(
  ae0,
  titles = c(
    "Table 14.3.1",
    "Overall Summary of Adverse Events",
    sprintf("Safety Population (N=%d)", saf["Total"])
  ),
  footnotes = "n (%) of subjects with at least one event in each category."
) |>
  cols(
    stat_label = col_spec(label = "Category"),
    placebo    = col_spec(label = sprintf("Placebo\n(N=%d)",  saf["placebo"]),  align = "decimal"),
    drug_50    = col_spec(label = sprintf("Drug 50\n(N=%d)",  saf["drug_50"]),  align = "decimal"),
    drug_100   = col_spec(label = sprintf("Drug 100\n(N=%d)", saf["drug_100"]), align = "decimal"),
    Total      = col_spec(label = sprintf("Total\n(N=%d)",    saf["Total"]),    align = "decimal")
  ) |>
  sort_rows(by = "Total", descending = TRUE)
```

## 3. Adverse events by system organ class and preferred term

The hierarchical AE table. Preferred terms indent under their
system-organ-class header via `indent_by`, and the columns that drive the
hierarchy and sort order are hidden. The data arrives in canonical
descending-frequency order, which `tabular` preserves.

```{r}
#| label: ae-socpt
tabular(
  saf_aesocpt,
  titles = c(
    "Table 14.3.2",
    "Adverse Events by System Organ Class and Preferred Term",
    sprintf("Safety Population (N=%d)", saf["Total"])
  ),
  footnotes = c(
    "Subjects are counted once within each system organ class and preferred term.",
    "MedDRA version 26.0."
  )
) |>
  cols(
    soc          = col_spec(visible = FALSE),
    label        = col_spec(usage = "group", indent_by = "indent_level",
                            label = "System Organ Class / Preferred Term"),
    row_type     = col_spec(visible = FALSE),
    indent_level = col_spec(visible = FALSE),
    n_total      = col_spec(visible = FALSE),
    soc_n        = col_spec(visible = FALSE),
    placebo      = col_spec(label = sprintf("Placebo\n(N=%d)",  saf["placebo"]),  align = "decimal"),
    drug_50      = col_spec(label = sprintf("Drug 50\n(N=%d)",  saf["drug_50"]),  align = "decimal"),
    drug_100     = col_spec(label = sprintf("Drug 100\n(N=%d)", saf["drug_100"]), align = "decimal"),
    Total        = col_spec(label = sprintf("Total\n(N=%d)",    saf["Total"]),    align = "decimal")
  ) |>
  headers("Treatment Group" = c("placebo", "drug_50", "drug_100", "Total"))
```

::: {.callout-tip}
## Keep a SOC with its terms in print

In a paginated backend, add `paginate(repeat_content = c("titles",
"headers", "footnotes"))` so the column band and footnotes repeat on
every page of this long table.
:::

## 4. Vital signs by parameter and visit

Each vital-sign parameter is a section; visit is an ordinary column so a
reader scans across timepoints within a parameter.

```{r}
#| label: vitals
tabular(
  saf_vital,
  titles = c(
    "Table 14.2.1",
    "Vital Signs by Parameter and Visit",
    sprintf("Safety Population (N=%d)", saf["Total"])
  ),
  footnotes = "Summary statistics of observed values at each visit."
) |>
  cols(
    paramcd    = col_spec(visible = FALSE),
    param      = col_spec(usage = "group"),
    visit      = col_spec(label = "Visit"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = sprintf("Placebo\n(N=%d)",  saf["placebo"]),  align = "decimal"),
    drug_50    = col_spec(label = sprintf("Drug 50\n(N=%d)",  saf["drug_50"]),  align = "decimal"),
    drug_100   = col_spec(label = sprintf("Drug 100\n(N=%d)", saf["drug_100"]), align = "decimal")
  ) |>
  headers("Treatment Group" = c("placebo", "drug_50", "drug_100"))
```

## 5. Best overall response and response rates

An efficacy table on the efficacy population: response categories grouped
under section labels (Best Overall Response, then derived rates), with the
efficacy BigN in the headers.

```{r}
#| label: efficacy
tabular(
  eff_resp,
  titles = c(
    "Table 14.2.1",
    "Best Overall Response and Response Rates",
    sprintf("Efficacy Population (N=%d)", sum(eff[c("placebo", "drug_50", "drug_100")]))
  ),
  footnotes = c(
    "Response assessed per RECIST 1.1.",
    "CI = confidence interval; ORR = objective response rate."
  )
) |>
  cols(
    row_type    = col_spec(visible = FALSE),
    groupid     = col_spec(visible = FALSE),
    group_label = col_spec(usage = "group"),
    stat_label  = col_spec(label = "Response"),
    placebo     = col_spec(label = sprintf("Placebo\n(N=%d)",  eff["placebo"]),  align = "decimal"),
    drug_50     = col_spec(label = sprintf("Drug 50\n(N=%d)",  eff["drug_50"]),  align = "decimal"),
    drug_100    = col_spec(label = sprintf("Drug 100\n(N=%d)", eff["drug_100"]), align = "decimal")
  ) |>
  headers("Treatment Group" = c("placebo", "drug_50", "drug_100"))
```

## 6. Subgrouped vital signs (Sex × Age group)

The same vitals analysis, partitioned into Sex × Age-group panels with
`subgroup()`. Each combination becomes its own banner-labelled partition
that starts a new page in the print backends.

```{r}
#| label: subgroup
tabular(
  saf_subgroup,
  titles = c(
    "Table 14.2.4",
    "Vital Signs by Sex and Age Group",
    "Safety Population"
  ),
  footnotes = "Summary statistics of observed values within each subgroup."
) |>
  cols(
    sex        = col_spec(visible = FALSE),
    agegr      = col_spec(visible = FALSE),
    sex_n      = col_spec(visible = FALSE),
    agegr_n    = col_spec(visible = FALSE),
    paramcd    = col_spec(visible = FALSE),
    param      = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  ) |>
  subgroup(by = c("sex", "agegr"), label = "{sex}, {agegr}")
```

## Shipping the deliverable

Every recipe ends the same way in production — pick a backend by file
extension and `emit()`:

```{r}
#| label: emit
#| eval: false
tab |> emit("t_14_1_1.rtf")     # RTF for the submission
tab |> emit("t_14_1_1.docx")    # native Word
tab |> emit("t_14_1_1.pdf")     # paginated PDF (via tinytex)
```

See [Fonts & fidelity](fonts-and-fidelity.html) for why the paginated
RTF/PDF/DOCX, not the HTML preview, is the artefact of record.
```
