---
title: "Data in: the cards → tabular pipeline"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

tabular displays a **wide** frame; clinical aggregation produces a **long**
Analysis Results Dataset (ARD). `pivot_across()` is the bridge — it widens an ARD
into the frame `tabular()` consumes.

![From analysis data to a display-ready frame: aggregate upstream into a long ARD, then `pivot_across()` widens it into the wide frame `tabular()` consumes.](../man/figures/data-pipeline.svg){fig-alt="An ADaM dataset is aggregated into a long ARD, which pivot_across widens into a wide data frame consumed by tabular." width="100%"}

The ARDs used below ship with the package, so these examples run without
cards/cardx installed; each is shown next to the code that produced it.

## The one rule: key `statistic` by the ARD's `context`

`pivot_across(statistic = list(...))` matches its list names against the ARD's
`context` column **verbatim**. That value depends on which function built the
ARD — use the wrong key and those rows are dropped **silently**. Always check
`unique(ard$context)` first.

| Generating function | `context` |
|---|---|
| `cards::ard_summary()` | `summary` |
| `cards::ard_tabulate()` | `tabulate` |
| `cards::ard_continuous()` | `continuous` |
| `cards::ard_categorical()` | `categorical` |
| `cards::ard_stack_hierarchical()` | `tabulate` + `hierarchical` |
| `cardx::ard_categorical_ci()` | `proportion_ci` |
| `cardx::ard_continuous_ci()` | `continuous_ci` |

A single string, or `statistic = list(default = ...)`, applies one format to
every context.

## Mixed continuous + categorical

`cdisc_saf_demo_ard` was stacked from a continuous and a categorical block, so it
carries the `continuous` and `categorical` contexts:

```{r eval = FALSE}
# how it was built (data-raw):
cdisc_saf_demo_ard <- cards::ard_stack(
  adsl,
  .by = TRT01A,
  cards::ard_continuous(
    variables = c(AGE, BMI),
    statistic = ~ cards::continuous_summary_fns(c(
      "N",
      "mean",
      "sd",
      "median",
      "min",
      "max"
    ))
  ),
  cards::ard_categorical(variables = c(AGEGR1, SEX, RACE))
)
```

```{r}
data(cdisc_saf_demo_ard, package = "tabular")
unique(cdisc_saf_demo_ard$context)

wide <- pivot_across(
  cdisc_saf_demo_ard,
  statistic = list(
    continuous = c(
      N = "{N}",
      "Mean (SD)" = "{mean} ({sd})",
      Median = "{median}",
      "Min, Max" = "{min}, {max}"
    ),
    categorical = "{n} ({p}%)"
  ),
  # CDISC precision: SD carries one more decimal than the mean. AGE/BMI are
  # whole numbers here (raw precision d = 0), so mean = d + 1 = 1, sd = d + 2 = 2,
  # median = d + 1 = 1, and min/max keep the raw precision (0).
  decimals = c(mean = 1, sd = 2, median = 1, min = 0, max = 0, p = 0)
)
head(wide, 8)
```

`decimals = c(p = 0)` gives integer percent with the pharma `<1` / `>99`
thresholds; `n = 0` cells collapse to a bare `"0"` automatically. The widened
frame is now ready for `tabular()` — see [Structure](structure.html).

## Confidence-interval cells (and hand-built ARDs)

cardx CI functions emit their own contexts (`proportion_ci`, `continuous_ci`)
with `estimate` / `conf.low` / `conf.high` stats. `pivot_across()` consumes **any
frame that follows the cards long-format**, so you can feed it a cardx ARD *or*
build the few rows by hand — useful when a statistic is computed outside cards:

```{r eval = FALSE}
# from cardx:
orr_ard <- cardx::ard_categorical_ci(
  resp,
  by = "TRT01A",
  variables = "RESP",
  method = "clopper-pearson",
  conf.level = 0.95
)
```

```{r}
# equivalent hand-built ARD (one row group, proportion_ci context).
# tribble() lays the long ARD out as a literal table: one row per
# (arm, stat), with the estimate and its CI bounds visible inline.
orr_ard <- tibble::tribble(
  ~group1,  ~group1_level,          ~variable, ~variable_level, ~context,        ~stat_name,  ~stat,
  "TRT01A", "Placebo",              "RESP",    "Responders",    "proportion_ci", "estimate",  0.62,
  "TRT01A", "Placebo",              "RESP",    "Responders",    "proportion_ci", "conf.low",  0.50,
  "TRT01A", "Placebo",              "RESP",    "Responders",    "proportion_ci", "conf.high", 0.73,
  "TRT01A", "Xanomeline Low Dose",  "RESP",    "Responders",    "proportion_ci", "estimate",  0.55,
  "TRT01A", "Xanomeline Low Dose",  "RESP",    "Responders",    "proportion_ci", "conf.low",  0.42,
  "TRT01A", "Xanomeline Low Dose",  "RESP",    "Responders",    "proportion_ci", "conf.high", 0.67,
  "TRT01A", "Xanomeline High Dose", "RESP",    "Responders",    "proportion_ci", "estimate",  0.48,
  "TRT01A", "Xanomeline High Dose", "RESP",    "Responders",    "proportion_ci", "conf.low",  0.36,
  "TRT01A", "Xanomeline High Dose", "RESP",    "Responders",    "proportion_ci", "conf.high", 0.60,
)

pivot_across(
  orr_ard,
  statistic = list(proportion_ci = "{estimate} ({conf.low}, {conf.high})"),
  decimals = c(estimate = 3, conf.low = 3, conf.high = 3)
)
```

## Hierarchical SOC / PT

A hierarchical ARD widens to a `soc` / `label` / `row_type` triple (not a single
`variable`), ready for an indented SOC ▸ PT layout:

```{r}
data(cdisc_saf_aesocpt_ard, package = "tabular")
ae <- pivot_across(cdisc_saf_aesocpt_ard, statistic = "{n} ({p}%)")
ae$indent_level <- as.integer(ae$row_type == "pt") # depth for col_spec(indent = "indent_level")
head(ae, 6)
```

(Turning `soc`/`label`/`indent_level` into the indented stub is in
[Structure](structure.html).)

## A two-variable `.by`

`ard_stack(.by = c(ARM, SEX))` carries a second grouping variable. Name it with
`row_group =` and `pivot_across()` widens it into a **leading row column**
(rather than mis-reading it as a SOC/PT hierarchy). Here is the cards long shape
such a stack produces (RACE by ARM within SEX), built by hand so the example runs
without cards:

```{r}
# RACE by ARM within SEX, laid out as a literal tribble: one row per
# (sex, race, arm, stat). Stats are placeholders (n = 10, p = 0.25).
two_by_ard <- tibble::tribble(
  ~group1, ~group1_level, ~group2, ~group2_level, ~variable, ~variable_level, ~context,      ~stat_name, ~stat,
  "ARM",   "Placebo",     "SEX",   "F",           "RACE",    "WHITE",         "categorical", "n",        10,
  "ARM",   "Placebo",     "SEX",   "F",           "RACE",    "WHITE",         "categorical", "p",        0.25,
  "ARM",   "Drug",        "SEX",   "F",           "RACE",    "WHITE",         "categorical", "n",        10,
  "ARM",   "Drug",        "SEX",   "F",           "RACE",    "WHITE",         "categorical", "p",        0.25,
  "ARM",   "Placebo",     "SEX",   "F",           "RACE",    "BLACK",         "categorical", "n",        10,
  "ARM",   "Placebo",     "SEX",   "F",           "RACE",    "BLACK",         "categorical", "p",        0.25,
  "ARM",   "Drug",        "SEX",   "F",           "RACE",    "BLACK",         "categorical", "n",        10,
  "ARM",   "Drug",        "SEX",   "F",           "RACE",    "BLACK",         "categorical", "p",        0.25,
  "ARM",   "Placebo",     "SEX",   "M",           "RACE",    "WHITE",         "categorical", "n",        10,
  "ARM",   "Placebo",     "SEX",   "M",           "RACE",    "WHITE",         "categorical", "p",        0.25,
  "ARM",   "Drug",        "SEX",   "M",           "RACE",    "WHITE",         "categorical", "n",        10,
  "ARM",   "Drug",        "SEX",   "M",           "RACE",    "WHITE",         "categorical", "p",        0.25,
  "ARM",   "Placebo",     "SEX",   "M",           "RACE",    "BLACK",         "categorical", "n",        10,
  "ARM",   "Placebo",     "SEX",   "M",           "RACE",    "BLACK",         "categorical", "p",        0.25,
  "ARM",   "Drug",        "SEX",   "M",           "RACE",    "BLACK",         "categorical", "n",        10,
  "ARM",   "Drug",        "SEX",   "M",           "RACE",    "BLACK",         "categorical", "p",        0.25,
)

pivot_across(
  two_by_ard,
  column = "ARM",
  row_group = "SEX",
  statistic = list(categorical = "{n} ({p}%)")
)
```

When a 2-variable `.by` is present you must say which variable is the arm column
(`column =`) and which is the row dimension (`row_group =`); the `SEX` column
then composes with `subgroup()` or `col_spec(usage = "group")` downstream. cards
encodes a crossing factor and a real hierarchy identically, so the declaration is
what disambiguates them — leave `row_group` unset for a genuine SOC/PT hierarchy.

## Variables as column bands

Two analysis variables side by side — `AVAL` ("Value") and `PCHG` ("Percent
Change from Baseline") — is the canonical "value and change" shell. Make the
**variable** a column band with the reserved `.variable` token in `column`. Each
band keys its own `statistic` / `decimals`, so the bands may carry different
(even different-length) stat lists; ragged bands pad with `NA`.

```{r}
# AVAL + PCHG by AVISIT x TRTA, built by hand so the example runs without cards.
valchg_ard <- tibble::tribble(
  ~group1,  ~group1_level, ~group2, ~group2_level, ~variable, ~context,      ~stat_name, ~stat,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "AVAL",    "continuous",  "N",        20,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "AVAL",    "continuous",  "mean",     324,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "AVAL",    "continuous",  "sd",       106,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "AVAL",    "continuous",  "median",   315,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "AVAL",    "continuous",  "N",        20,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "AVAL",    "continuous",  "mean",     318,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "AVAL",    "continuous",  "sd",       98,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "AVAL",    "continuous",  "median",   310,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "PCHG",    "continuous",  "N",        20,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "PCHG",    "continuous",  "mean",     -16,
  "AVISIT", "DAY 1",       "TRTA",  "Drug",        "PCHG",    "continuous",  "sd",       5,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "PCHG",    "continuous",  "N",        20,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "PCHG",    "continuous",  "mean",     -14,
  "AVISIT", "DAY 1",       "TRTA",  "Placebo",     "PCHG",    "continuous",  "sd",       5,
)
```

**Stats as rows** — `column = c(".variable", "<arm>")`. Each variable becomes a
band of arm columns; the statistics stack as rows and cells are combined strings.
The emitted columns are named `"<variable>..<arm>"`:

```{r}
pivot_across(
  valchg_ard,
  column = c(".variable", "TRTA"),
  row_group = "AVISIT",
  statistic = list(
    AVAL = c(N = "{N}", "Mean (SD)" = "{mean} ({sd})", Median = "{median}"),
    PCHG = c(N = "{N}", "Mean (SD)" = "{mean} ({sd})")
  ),
  decimals = list(AVAL = c(mean = 1, sd = 2, median = 1), PCHG = c(mean = 1, sd = 2))
)
```

**Stats as columns** — `column = c(".variable", ".stat")`. Each statistic entry
becomes its own column (the landscape shell) and the arm drops to a leading row
stub. The emitted columns are named `"<variable>..<stat-entry>"`:

```{r}
pivot_across(
  valchg_ard,
  column = c(".variable", ".stat"),
  row_group = "AVISIT",
  statistic = list(
    AVAL = c(N = "{N}", Mean = "{mean}", SD = "{sd}"),
    PCHG = c(N = "{N}", Mean = "{mean}")
  ),
  decimals = c(mean = 1, sd = 2)
)
```

You reference the emitted `"<variable>..<arm>"` / `"<variable>..<stat>"` names
verbatim in a manual `headers()` call to draw the band spanners — `pivot_across()`
never builds spanners itself.

## Auxiliary comparison columns

A between-arm comparison (difference, hazard ratio, p-value) is **not** a pivot of
the main ARD's rows — it is a separate ARD (e.g. from `cardx`). Bind it with
`aux =`, aligned 1:1 on the `row_group` key; the entry name becomes the column.

```{r}
resp_main <- tibble::tribble(
  ~group1,  ~group1_level, ~group2, ~group2_level, ~variable, ~context,     ~stat_name, ~stat,
  "PARAM",  "ORR",         "TRTA",  "Exp",         "AVAL",    "continuous", "mean",     2.2,
  "PARAM",  "ORR",         "TRTA",  "Ctl",         "AVAL",    "continuous", "mean",     1.9,
  "PARAM",  "DCR",         "TRTA",  "Exp",         "AVAL",    "continuous", "mean",     2.4,
  "PARAM",  "DCR",         "TRTA",  "Ctl",         "AVAL",    "continuous", "mean",     2.0,
)
diff_ard <- tibble::tribble(
  ~group1, ~group1_level, ~variable, ~context,     ~stat_name, ~stat,
  "PARAM", "ORR",         "d",       "continuous", "mean",     0.12,
  "PARAM", "DCR",         "d",       "continuous", "mean",     0.20,
)

pivot_across(
  resp_main,
  column = "TRTA",
  row_group = "PARAM",
  statistic = list(continuous = "{mean}"),
  aux = list(
    "Difference" = list(ard = diff_ard, statistic = "{mean}", decimals = c(mean = 2))
  )
)
```

One `aux` entry is one column; add more entries (estimate then p-value) for
several comparison columns. The auxiliary ARD must reduce to one row per
`row_group` key — a many-to-many alignment aborts rather than fabricate rows.

## Key `statistic` by the context

The one thing to get right: key `statistic` by the ARD's `context`. Inspect
`unique(ard$context)` and key the list to match (or pass a single string /
`default =`). If an explicitly-supplied `statistic` matches **no** context,
`pivot_across()` warns rather than silently falling back to `{n}`.
