---
title: "Demo datasets"
description: >
  A tour of the eleven bundled datasets: what each represents, its shape,
  and a rendered preview. Use them to learn, prototype, and test.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

Eleven pre-summarised datasets ship with `tabular`. They power every
example, vignette, and test, and they are the fastest way to prototype a
table without preparing your own data. All are synthetic, derived from
the CDISC pilot compounds (Xanomeline and placebo).

## Safety tables

### `saf_demo` — demographics

The canonical demographics summary: one row per statistic per
characteristic, columns per arm. The `variable` column is the section
grouping; `stat_label` the statistic.

```{r}
#| label: saf_demo
dim(saf_demo)
names(saf_demo)

tabular(saf_demo[1:7, ]) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

### `saf_aeoverall` — high-level AE summary

A compact overall adverse-event table: one row per AE category, columns
per arm.

```{r}
#| label: saf_aeoverall
dim(saf_aeoverall)

tabular(saf_aeoverall[c("stat_label", "placebo", "drug_50", "drug_100", "Total")]) |>
  cols(
    stat_label = col_spec(label = "Category"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

### `saf_aesocpt` — AEs by SOC and preferred term

A hierarchical AE table. `label` carries both the system-organ-class and
preferred-term text; `indent_level`, `row_type`, and the sort keys
(`soc_n`, `n_total`) are hidden columns that drive the display.

```{r}
#| label: saf_aesocpt
dim(saf_aesocpt)
names(saf_aesocpt)
```

### `saf_vital` — vital signs by parameter and visit

Four parameters × four visits × summary statistics. `param` is the
section; `visit` an ordinary column.

```{r}
#| label: saf_vital
dim(saf_vital)
levels(factor(saf_vital$visit))
```

### `saf_subgroup` — vitals by Sex × Age group

The subgrouped vitals analysis. `sex` and `agegr` partition the table
(with per-partition BigN in `sex_n` / `agegr_n`); use them with
`subgroup()`.

```{r}
#| label: saf_subgroup
dim(saf_subgroup)
names(saf_subgroup)
```

## Efficacy tables

### `eff_resp` — best overall response and rates

Response categories grouped under section labels (`group_label`): Best
Overall Response, then derived Objective Response / Clinical Benefit /
Disease Control rates.

```{r}
#| label: eff_resp
dim(eff_resp)
unique(eff_resp$group_label)
```

### `eff_estimates` — treatment-effect estimates

Raw **numeric** estimates (not pre-formatted strings) from several
models — a fixture for exercising `col_spec(format =)` and numeric
rounding at render time.

```{r}
#| label: eff_estimates
eff_estimates
```

## Bridge and denominator data

### `saf_demo_card`, `saf_aesocpt_card` — cards ARDs

Long-format Analysis Results Data in the `cards` shape, the input to
`pivot_across()`. `saf_aesocpt_card` is hierarchical (SOC within PT). See
[From a cards ARD](from-ard.html).

```{r}
#| label: cards
dim(saf_demo_card)
head(saf_demo_card, 4)
```

### `saf_n`, `eff_n` — BigN denominators

Per-arm subject counts for the safety and efficacy populations. Key the
column-header `(N=…)` labels off `arm_short` rather than carrying the
counts as a fragile attribute.

```{r}
#| label: bign
saf_n
eff_n
```

## Quick reference

| Dataset | Rows × Cols | Represents |
|---|---|---|
| `saf_demo` | 35 × 6 | Demographics, Safety Population |
| `saf_aeoverall` | 8 × 5 | Overall AE summary |
| `saf_aesocpt` | 61 × 10 | AEs by SOC / preferred term |
| `saf_vital` | 64 × 7 | Vital signs by parameter and visit |
| `saf_subgroup` | 32 × 11 | Vitals by Sex × Age group |
| `eff_resp` | 13 × 7 | Best overall response and rates |
| `eff_estimates` | 4 × 5 | Treatment-effect estimates (numeric) |
| `saf_demo_card` | 317 × 8 | Demographics ARD (long) |
| `saf_aesocpt_card` | 558 × 10 | AE ARD, hierarchical (long) |
| `saf_n` | 4 × 3 | Safety BigN per arm |
| `eff_n` | 4 × 3 | Efficacy BigN per arm |
```
