---
title: "Data in: the cards → tabular pipeline"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

tabular displays a **wide** frame; clinical aggregation produces a **long**
Analysis Results Dataset (ARD). `pivot_across()` is the bridge — it widens an ARD
into the frame `tabular()` consumes.

![From analysis data to a display-ready frame: aggregate upstream into a long ARD, then `pivot_across()` widens it into the wide frame `tabular()` consumes.](../man/figures/data-pipeline.svg){fig-alt="An ADaM dataset is aggregated into a long ARD, which pivot_across widens into a wide data frame consumed by tabular." width="100%"}

The ARDs used below ship with the package, so these examples run without
cards/cardx installed; each is shown next to the code that produced it.

## The one rule: key `statistic` by the ARD's `context`

`pivot_across(statistic = list(...))` matches its list names against the ARD's
`context` column **verbatim**. That value depends on which function built the
ARD — use the wrong key and those rows are dropped **silently**. Always check
`unique(ard$context)` first.

| Generating function | `context` |
|---|---|
| `cards::ard_summary()` | `summary` |
| `cards::ard_tabulate()` | `tabulate` |
| `cards::ard_continuous()` | `continuous` |
| `cards::ard_categorical()` | `categorical` |
| `cards::ard_stack_hierarchical()` | `tabulate` + `hierarchical` |
| `cardx::ard_categorical_ci()` | `proportion_ci` |
| `cardx::ard_continuous_ci()` | `continuous_ci` |

A single string, or `statistic = list(default = ...)`, applies one format to
every context.

## Mixed continuous + categorical

`cdisc_saf_demo_ard` was stacked from a continuous and a categorical block, so it
carries the `continuous` and `categorical` contexts:

```{r eval = FALSE}
# how it was built (data-raw):
cdisc_saf_demo_ard <- cards::ard_stack(
  adsl,
  .by = TRT01A,
  cards::ard_continuous(
    variables = c(AGE, BMI),
    statistic = ~ cards::continuous_summary_fns(c(
      "N",
      "mean",
      "sd",
      "median",
      "min",
      "max"
    ))
  ),
  cards::ard_categorical(variables = c(AGEGR1, SEX, RACE))
)
```

```{r}
data(cdisc_saf_demo_ard, package = "tabular")
unique(cdisc_saf_demo_ard$context)

wide <- pivot_across(
  cdisc_saf_demo_ard,
  statistic = list(
    continuous = c(
      N = "{N}",
      "Mean (SD)" = "{mean} ({sd})",
      Median = "{median}",
      "Min, Max" = "{min}, {max}"
    ),
    categorical = "{n} ({p}%)"
  ),
  # CDISC precision: SD carries one more decimal than the mean. AGE/BMI are
  # whole numbers here (raw precision d = 0), so mean = d + 1 = 1, sd = d + 2 = 2,
  # median = d + 1 = 1, and min/max keep the raw precision (0).
  decimals = c(mean = 1, sd = 2, median = 1, min = 0, max = 0, p = 0)
)
head(wide, 8)
```

`decimals = c(p = 0)` gives integer percent with the pharma `<1` / `>99`
thresholds; `n = 0` cells collapse to a bare `"0"` automatically. The widened
frame is now ready for `tabular()` — see [Structure](structure.html).

## Confidence-interval cells (and hand-built ARDs)

cardx CI functions emit their own contexts (`proportion_ci`, `continuous_ci`)
with `estimate` / `conf.low` / `conf.high` stats. `pivot_across()` consumes **any
frame that follows the cards long-format**, so you can feed it a cardx ARD *or*
build the few rows by hand — useful when a statistic is computed outside cards:

```{r eval = FALSE}
# from cardx:
orr_ard <- cardx::ard_categorical_ci(
  resp,
  by = "TRT01A",
  variables = "RESP",
  method = "clopper-pearson",
  conf.level = 0.95
)
```

```{r}
# equivalent hand-built ARD (one row group, proportion_ci context)
arms <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
orr_ard <- expand.grid(
  group1_level = arms,
  stat_name = c("estimate", "conf.low", "conf.high"),
  stringsAsFactors = FALSE
)
orr_ard$group1 <- "TRT01A"
orr_ard$variable <- "RESP"
orr_ard$variable_level <- "Responders"
orr_ard$context <- "proportion_ci"
orr_ard$stat <- c(0.62, 0.55, 0.48, 0.50, 0.42, 0.36, 0.73, 0.67, 0.60)

pivot_across(
  orr_ard,
  statistic = list(proportion_ci = "{estimate} ({conf.low}, {conf.high})"),
  decimals = c(estimate = 3, conf.low = 3, conf.high = 3)
)
```

## Hierarchical SOC / PT

A hierarchical ARD widens to a `soc` / `label` / `row_type` triple (not a single
`variable`), ready for an indented SOC ▸ PT layout:

```{r}
data(cdisc_saf_aesocpt_ard, package = "tabular")
ae <- pivot_across(cdisc_saf_aesocpt_ard, statistic = "{n} ({p}%)")
ae$indent <- as.integer(ae$row_type == "pt") # depth for col_spec(indent_by=)
head(ae, 6)
```

(Turning `soc`/`label`/`indent` into the indented stub is in
[Structure](structure.html).)

## A two-variable `.by`

`ard_stack(.by = c(ARM, SEX))` carries a second grouping variable. Name it with
`row_group =` and `pivot_across()` widens it into a **leading row column**
(rather than mis-reading it as a SOC/PT hierarchy). Here is the cards long shape
such a stack produces (RACE by ARM within SEX), built by hand so the example runs
without cards:

```{r}
two_by_ard <- expand.grid(
  group1_level = c("Placebo", "Drug"),
  group2_level = c("F", "M"),
  variable_level = c("WHITE", "BLACK"),
  stat_name = c("n", "p"),
  stringsAsFactors = FALSE
)
two_by_ard$group1 <- "ARM"
two_by_ard$group2 <- "SEX"
two_by_ard$variable <- "RACE"
two_by_ard$context <- "categorical"
two_by_ard$stat <- ifelse(two_by_ard$stat_name == "n", 10, 0.25)

pivot_across(
  two_by_ard,
  column = "ARM",
  row_group = "SEX",
  statistic = list(categorical = "{n} ({p}%)")
)
```

When a 2-variable `.by` is present you must say which variable is the arm column
(`column =`) and which is the row dimension (`row_group =`); the `SEX` column
then composes with `subgroup()` or `col_spec(usage = "group")` downstream. cards
encodes a crossing factor and a real hierarchy identically, so the declaration is
what disambiguates them — leave `row_group` unset for a genuine SOC/PT hierarchy.

## Key `statistic` by the context

The one thing to get right: key `statistic` by the ARD's `context`. Inspect
`unique(ard$context)` and key the list to match (or pass a single string /
`default =`). If an explicitly-supplied `statistic` matches **no** context,
`pivot_across()` warns rather than silently falling back to `{n}`.
