---
title: "Fonts & fidelity"
description: >
  Why clinical tables are monospace, how decimal alignment depends on
  font metrics, auditing fonts with check_fonts(), and why the print
  backends, not the HTML preview, are the artefact of record.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

A submission table is a *typeset* artefact. Two facts follow from that,
and they shape how `tabular` renders: clinical tables are monospace, and
the deliverable is the paginated print output, not the screen preview.

## Why monospace

Reviewers scan a column of numbers by eye, so the digits must line up on
the decimal point. That only works reliably when every character has the
**same advance width** — a monospace font. In a proportional font, `1`
is narrower than `8`, and a column of `11.1` over `88.8` drifts out of
alignment.

`tabular` therefore renders table cells in a monospace family and aligns
`align = "decimal"` columns using the backend's **real font metrics**, not
guessed spaces. The alignment computed at Resolve time is the alignment
you get on the page.

```{r}
#| label: decimal-demo
tabular(saf_aeoverall) |>
  cols(
    stat_label = col_spec(label = "Category"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  )
```

## Choosing a font

Regulatory tables are conventionally set in **Courier New** or **Times
New Roman** at 8–9 pt; agency style guidance (FDA, EMA, PMDA) drives the
exact choice per submission. Set the family and size on `preset()`:

```{r}
#| label: font-preset
tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(label = "Placebo",  align = "decimal"),
    drug_50    = col_spec(label = "Drug 50",  align = "decimal"),
    drug_100   = col_spec(label = "Drug 100", align = "decimal"),
    Total      = col_spec(label = "Total",    align = "decimal")
  ) |>
  preset(font_family = "Courier New", font_size = 9)
```

## Auditing availability with `check_fonts()`

Each backend resolves a font through a fallback stack. `check_fonts()`
reports which fonts in the active stack are actually installed on the
machine and returns the per-backend stacks, so you can confirm the
submission font is present before a batch run:

```{r}
#| label: check-fonts
sp <- tabular(saf_demo) |>
  cols(
    variable   = col_spec(usage = "group"),
    stat_label = col_spec(label = "Statistic"),
    placebo    = col_spec(align = "decimal"),
    drug_50    = col_spec(align = "decimal"),
    drug_100   = col_spec(align = "decimal"),
    Total      = col_spec(align = "decimal")
  )

stacks <- check_fonts(sp)
stacks$rtf
```

A missing font is not fatal — the backend falls through the stack — but
for a submission you want the named font present so the rendered metrics
match the agency's expectation.

## The preview is not the deliverable

This site shows a **live HTML preview** of every table, which is
invaluable while you build. But the artefact you ship is the paginated
**RTF, PDF, or DOCX**:

| | HTML preview | RTF / PDF / DOCX deliverable |
|---|---|---|
| Pages | one continuous document | real, paginated |
| Repeated chrome | n/a | titles / headers / footnotes per page |
| Font metrics | browser CSS monospace | backend's real font metrics |
| Page layout | scrollable | the four-section submission layout |
| Role | build & eyeball | the artefact of record |

Both come from the **same resolved grid**, so the numbers and structure
cannot drift between them — but pagination, repeated chrome, and exact
metrics only exist in the print backends. Build and check in HTML, then:

```{r}
#| label: emit
#| eval: false
sp |> emit("t_14_1_1.rtf")     # the deliverable
sp |> emit("t_14_1_1.pdf")     # paginated PDF via tinytex
sp |> emit("t_14_1_1.docx")    # native Word
```

## Where to next

- **[Architecture](architecture.html)** — how Resolve computes the
  metrics-aware grid that every backend consumes.
- **[Clinical cookbook](clinical-cookbook.html)** — the recipes you would
  emit to these formats.
```
