---
title: "From a cards ARD"
description: >
  Bridge long-format Analysis Results Data into the wide frame tabular
  renders, with pivot_across().
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

`tabular` renders a **wide** frame, but many analysis pipelines produce
**long** Analysis Results Data (ARD) — one row per statistic, in the
`cards` format. `pivot_across()` is the bridge: it pivots an ARD into the
wide, one-row-per-display-row shape `tabular()` expects.

::: {.callout-note}
## What is an ARD?

An *Analysis Results Data* frame is the tidy, long representation of a
table's contents: each row carries one statistic for one group level, with
columns like `group1_level`, `variable`, `stat_name`, and `stat`. It is
produced upstream by `cards::ard_stack()` (and friends). `tabular` does
not compute it — it consumes it.
:::

## A demographics ARD

The bundled `saf_demo_card` is a demographics ARD in `cards` format:

```{r}
#| label: ard-peek
head(saf_demo_card, 6)
```

Each row is one statistic (`stat`) for one treatment arm
(`group1_level`) and characteristic (`variable`). That is too long to
render directly.

## Pivot to wide

`pivot_across()` spreads the treatment arms across columns and formats
each statistic into a display string:

```{r}
#| label: pivot
wide <- pivot_across(saf_demo_card)

wide
```

By default the `statistic` argument formats continuous rows as
`"{mean} ({sd})"` and categorical rows as `"{n} ({p}%)"`, and an
`overall = "Total"` column is appended. The arm columns take their names
from the ARD's `group1_level` values.

## Render the pivoted frame

The result is an ordinary wide data frame, so it flows straight into the
usual pipeline. Arm columns whose names contain spaces are quoted with
backticks:

```{r}
#| label: render
tabular(
  wide,
  titles = c(
    "Table 14.1.1",
    "Demographic and Baseline Characteristics",
    "Safety Population"
  ),
  footnotes = "Built from a cards ARD via pivot_across()."
) |>
  cols(
    variable               = col_spec(usage = "group", label = "Characteristic"),
    stat_label             = col_spec(label = "Statistic"),
    Placebo                = col_spec(align = "decimal"),
    `Xanomeline Low Dose`  = col_spec(label = "Drug 50",  align = "decimal"),
    `Xanomeline High Dose` = col_spec(label = "Drug 100", align = "decimal"),
    Total                  = col_spec(align = "decimal")
  )
```

## Customising the statistic format

Pass your own templates to `statistic` to change how each context renders
— for example, a mean with one extra decimal and percentages without
parentheses:

```{r}
#| label: custom-stat
pivot_across(
  saf_demo_card,
  statistic = list(
    continuous  = "{mean} (SD {sd})",
    categorical = "{n} / {p}%"
  )
)
```

The `column` argument selects which ARD grouping variable becomes the
columns (handy when an ARD carries more than one grouping dimension), and
`label` renames the `variable` values for display.

::: {.callout-tip}
## Hierarchical ARDs

`saf_aesocpt_card` is a *hierarchical* AE ARD (system organ class within
preferred term). The same `pivot_across()` entry point handles it; the
resulting wide frame then renders with the indent-and-group recipe from
the [Clinical cookbook](clinical-cookbook.html).
:::

## Where this fits

`pivot_across()` keeps the division of labour clean: `cards` (or
`gtsummary`, `dplyr`, SAS) *computes* the summary; `pivot_across()`
reshapes it; `tabular()` *renders* it to submission-grade output. See the
[Comparison article](comparison.html) for how that split compares to
all-in-one table packages.
```
