---
title: "From a cards ARD"
description: >
  Bridge long-format Analysis Results Data (the cards ARD) into the wide
  frame tabular renders, with pivot_across() -- demographics and a
  hierarchical adverse-event table, end to end.
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| label: setup
library(tabular)
```

`tabular` renders a **wide** frame -- one row per display row, one column
per arm. But most modern analysis pipelines emit **long** Analysis
Results Data (an *ARD*) in the `cards` format: one row per statistic.
`pivot_across()` is the bridge between the two. This article walks a
demographics ARD and a hierarchical adverse-event ARD all the way from
their long form to a rendered table.

::: {.callout-note}
## What is an ARD?

An *Analysis Results Data* frame is the tidy, long representation of a
table's contents: each row carries one statistic, for one group level, of
one variable. It is produced upstream by `cards::ard_stack()` and
friends. `tabular` does not compute it -- it consumes it. The division of
labour is deliberate: `cards` (or `gtsummary`, `dplyr`, SAS) *computes*
the numbers; `pivot_across()` *reshapes* them; `tabular()` *renders*.
:::

## 1. Inspect the ARD

The bundled `saf_demo_card` is a demographics ARD. One row is one
statistic, so it is far too long to render directly -- the first six rows
cover just the start of one arm's age summary:

```{r}
#| label: ard-peek
head(saf_demo_card, 6)
```

The columns that matter to `pivot_across()` are:

- **`group1_level`** -- the treatment arm (`Placebo`, `Xanomeline Low
  Dose`, ...). These become the table's columns.
- **`variable`** -- the characteristic (`AGE`, `SEX`, ...). These become
  the row sections.
- **`context`** -- `continuous` or `categorical`, which selects the
  formatting template.
- **`stat_name` / `stat_label` / `stat`** -- the statistic's code, its
  display name (`Mean`, `SD`, `n`, `%`), and its raw numeric value.

## 2. Pivot to wide

`pivot_across()` spreads `group1_level` across columns and formats each
statistic into one display string:

```{r}
#| label: pivot
wide <- pivot_across(saf_demo_card)
head(wide, 8)
```

By default continuous rows format as `"{mean} ({sd})"`, categorical rows
as `"{n} ({p}%)"`, and an `overall = "Total"` column is appended. The arm
columns take their names verbatim from `group1_level`, so spaces survive
(`Xanomeline Low Dose`) and need backticks when you name them in
`cols()`.

## 3. Render it

The pivot leaves `variable` as the raw upstream codes (`AGE`, `SEX`). Pass
a `label` map to `pivot_across()` to rename them to display labels in one
place -- it rewrites the `variable` (and, for hierarchical ARDs, the `soc`
/ `label`) values. The wide frame is then an ordinary `tabular()` input:

```{r}
#| label: render
wide <- pivot_across(
  saf_demo_card,
  label = c(
    AGE     = "Age (years)",
    WEIGHT  = "Weight (kg)",
    HEIGHT  = "Height (cm)",
    BMI     = "BMI (kg/m²)",
    AGEGR1  = "Age Group, n (%)",
    SEX     = "Sex, n (%)",
    RACE    = "Race, n (%)",
    ETHNIC  = "Ethnicity, n (%)",
    BMI_CAT = "BMI Category, n (%)"
  )
)

tabular(
  wide,
  titles = c(
    "Table 14.1.1",
    "Demographic and Baseline Characteristics",
    "Safety Population"
  ),
  footnotes = "Built from a cards ARD via pivot_across()."
) |>
  cols(
    variable               = col_spec(usage = "group", label = "Characteristic"),
    stat_label             = col_spec(label = "Statistic"),
    Placebo                = col_spec(label = "Placebo",  align = "decimal"),
    `Xanomeline Low Dose`  = col_spec(label = "Drug 50",  align = "decimal"),
    `Xanomeline High Dose` = col_spec(label = "Drug 100", align = "decimal"),
    Total                  = col_spec(label = "Total",    align = "decimal")
  )
```

`usage = "group"` on `variable` turns each characteristic into a bold
section-header row; `stat_label` carries the per-row statistic name.

## 4. Change the statistic format

The `statistic` argument takes one template per context. Swap in your own
to change how every row of that kind renders -- here a mean carrying its
SD spelled out, and percentages without parentheses. Because the result
is still a wide frame, it renders the same way:

```{r}
#| label: custom-stat
pivot_across(
  saf_demo_card,
  statistic = list(
    continuous  = "{mean} (SD {sd})",
    categorical = "{n} / {p}%"
  ),
  label = c(AGE = "Age (years)", SEX = "Sex, n (%)")
) |>
  tabular(titles = "Demographics with a custom statistic format") |>
  cols(
    variable               = col_spec(usage = "group", label = "Characteristic"),
    stat_label             = col_spec(label = "Statistic"),
    Placebo                = col_spec(label = "Placebo",  align = "decimal"),
    `Xanomeline Low Dose`  = col_spec(label = "Drug 50",  align = "decimal"),
    `Xanomeline High Dose` = col_spec(label = "Drug 100", align = "decimal"),
    Total                  = col_spec(label = "Total",    align = "decimal")
  )
```

`decimals` and `fmt` give finer control over rounding when a template
alone is not enough; `column` selects which grouping variable becomes the
columns when an ARD carries more than one.

### Integer percentages

`decimals` is a named vector keyed by statistic. Some sponsor shells show
percentages as whole numbers — pass `decimals = c(p = 0)` to drop the
decimal place on the `{p}` stat while leaving the others untouched:

```{r}
#| label: integer-pct
pivot_across(
  saf_demo_card,
  decimals = c(p = 0),
  label = c(SEX = "Sex, n (%)", RACE = "Race, n (%)")
) |>
  tabular(titles = "Demographics with integer percentages") |>
  cols(
    variable               = col_spec(usage = "group", label = "Characteristic"),
    stat_label             = col_spec(label = "Statistic"),
    Placebo                = col_spec(label = "Placebo",  align = "decimal"),
    `Xanomeline Low Dose`  = col_spec(label = "Drug 50",  align = "decimal"),
    `Xanomeline High Dose` = col_spec(label = "Drug 100", align = "decimal"),
    Total                  = col_spec(label = "Total",    align = "decimal")
  )
```

The pharma rounding threshold tracks the precision you ask for: at
`decimals = c(p = 0)` a non-zero percentage that would round to `0`
renders as `<1`, and one that would round to `100` renders as `>99`, so a
rare event never disappears into `0%`. Raise the precision and the
threshold follows — `decimals = c(p = 2)` gives `<0.01` / `>99.99`.

## 5. A hierarchical AE ARD

`saf_aesocpt_card` is a *hierarchical* AE ARD -- preferred terms nested
inside system organ classes. `pivot_across()` flattens it into a wide
frame carrying `soc`, a combined `label`, and a `row_type`
(`overall` / `soc` / `pt`):

```{r}
#| label: ae-pivot
ae <- pivot_across(saf_aesocpt_card)
head(ae, 4)
```

Two small touches turn that into a submission-shaped table. The pivot
does not ship an indent key, so derive one from `row_type`; and the
all-subjects row arrives with an internal placeholder label, so give it a
real one:

```{r}
#| label: ae-prep
ae$indent_level <- ifelse(ae$row_type == "pt", 1L, 0L)
ae$label[ae$row_type == "overall"] <- "Total subjects with an event"
```

Now it renders with the same indent-and-bold recipe as the
[Clinical cookbook](clinical-cookbook.html): the `soc` / `row_type`
columns ride along hidden, the single `label` column indents preferred
terms beneath their SOC via `indent_by`, and a `style()` layer bolds the
overall and SOC summary rows:

```{r}
#| label: ae-render
tabular(
  ae,
  titles = c(
    "Table 14.3.1",
    "Adverse Events by System Organ Class and Preferred Term",
    "Safety Population"
  ),
  footnotes = "Built from a hierarchical cards ARD via pivot_across()."
) |>
  cols(
    soc                    = col_spec(visible = FALSE),
    row_type               = col_spec(visible = FALSE),
    indent_level           = col_spec(visible = FALSE),
    label                  = col_spec(label = "System Organ Class / Preferred Term",
                                      indent_by = "indent_level"),
    Placebo                = col_spec(label = "Placebo",  align = "decimal"),
    `Xanomeline Low Dose`  = col_spec(label = "Drug 50",  align = "decimal"),
    `Xanomeline High Dose` = col_spec(label = "Drug 100", align = "decimal")
  ) |>
  style(bold = TRUE, .at = cells_body(where = row_type %in% c("overall", "soc")))
```

## Where this fits

`pivot_across()` keeps the division of labour clean: an upstream tool
*computes* the summary as an ARD, `pivot_across()` *reshapes* it to the
wide frame, and `tabular()` *renders* it to submission-grade RTF / PDF /
DOCX / HTML. Nothing in the rendering stack needs to know how the numbers
were produced -- only their shape.

## Where to next

- **[Clinical cookbook](clinical-cookbook.html)** -- six complete
  production tables, including the AE and demographics recipes this
  article feeds.
- **[Columns & headers](columns-and-headers.html)** -- every `col_spec()`
  option used above, in depth.
