---
title: "Output & qualification: backends, requirements, and the CDISC pilot"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

This article is about *rendering and proving* — choosing a backend, meeting its
system requirements, and the cross-backend validation. It does not cover building
or styling a table (see the other articles).

## `emit()` and `as_grid()`

`emit()` writes a file, dispatching on the extension:

```{r eval = FALSE}
emit(spec, "table.rtf") # RTF
emit(spec, "table.html") # HTML
emit(spec, "table.docx") # Word
emit(spec, "table.tex") # LaTeX source
emit(spec, "table.typ") # Typst source
emit(spec, "table.pdf") # PDF (LaTeX engine, or typst without a TeX)
emit(spec, "table.md") # Markdown
```

A `.pdf` target compiles through one of **two engines**. With no
`format =`, `emit()` probes the machine LaTeX-first: a usable TeX keeps
the LaTeX path; otherwise a discoverable typst binary (standalone
`typst`, or the copy bundled inside Quarto ≥ 1.4) takes over; with
neither, the call aborts up front naming both remedies. Pick an engine
explicitly with `format = "latex"` or `format = "typst"`:

```{r eval = FALSE}
emit(spec, "table.pdf", format = "latex") # force the TeX compile
emit(spec, "table.pdf", format = "typst") # force typst — no TeX needed
```

It returns the written path invisibly, so the emit is chainable into
scripted batch runs. One spec, any backend:

```{r}
data(cdisc_saf_demo, package = "tabular")
spec <- tabular(cdisc_saf_demo, titles = "Demographics") |>
  cols(
    variable = col_spec(label = ""),
    stat_label = col_spec(label = "")
  ) |>
  group_rows(by = "variable")

path <- emit(spec, tempfile(fileext = ".rtf"))
file.exists(path)
```

`as_grid(spec)` resolves the fully-laid-out grid **without** writing a file —
useful for testing or programmatic inspection. The grid carries the resolved
pages plus a metadata block (pagination counts, resolved column names, the
effective preset):

```{r}
grid <- as_grid(spec)
length(grid@pages)
grid@metadata$col_names
```

## Backend capability matrix

One spec renders to every backend, but the page-oriented features differ:

| Capability | RTF | HTML | DOCX | PDF/LaTeX | PDF/Typst | MD |
|---|:--:|:--:|:--:|:--:|:--:|:--:|
| Vertical pagination | ✓ | n/a¹ | ✓ | ✓ | ✓ | n/a |
| Horizontal panels (`panels=`) | ✓ | n/a¹ | ✓ | ✓ | ✓ | n/a |
| Per-page running header/footer | ✓ | – | ✓ | ✓ | ✓ | – |
| `subgroup()` per-page BigN | ✓ | row² | ✓ | ✓ | ✓ | row² |
| Continuation marker | panels only | – | – | ✓ | panels only | – |
| Keep-together / orphan control | ✓ | n/a¹ | ✓ | ✓ | ✓ | n/a |
| Decimal alignment (NBSP) | ✓ | ✓ | ✓ | ✓ | ✓ | ✓ |
| System dependency | none | none | none | TeX install | typst³ | none |

¹ HTML/MD are one continuous document; the browser repeats `<thead>` on print.

² On HTML/MD the per-page N renders as a row under each subgroup banner instead
of in the repeating header.

³ The standalone `typst` binary, or the copy bundled inside Quarto ≥ 1.4 —
so machines with RStudio / Posit Workbench typically need nothing. The `.tex`
and `.typ` *source* backends have no system dependency at all; only the
compile to PDF does.

## System requirements

**RTF, HTML, DOCX, LaTeX source, Typst source, Markdown need nothing beyond
the R package.** Only the compile to PDF has a system dependency, and either
engine satisfies it:

- **LaTeX engine.** The two packages missing from common TeX distributions
  (`tabularray`, `ninecolors`) ship *with* `tabular` and are staged next to
  the generated `.tex` whenever the local TeX cannot resolve them, so a
  locked-down server (Domino, Posit Workbench) needs no `tlmgr install`.
- **Typst engine.** No TeX at all: `emit()` runs the standalone `typst`
  binary, or Quarto's bundled copy (`quarto typst`). Fonts are typst's one
  quiet failure mode — a missing family substitutes silently — so `emit()`
  warns when a family you explicitly named cannot be found, and
  `check_typst()` names the face PDFs actually render in.

Check readiness and, on a fresh machine, install an engine once:

```{r eval = FALSE}
check_latex() # LaTeX engine: probes via kpsewhich — what a compile will find
check_typst() # Typst engine: binary, version floor, font chain
tinytex::install_tinytex(bundle = "TinyTeX") # one-time, fresh machines only
# or install Quarto (https://quarto.org), which bundles the typst engine
```

> **OS-managed TeX Live (RHEL/dnf, Debian/apt):** `tlmgr` is locked and refuses
> to install ("will likely destroy the … TeXLive install"). Do **not** force it
> with `--ignore-warning`. Usually nothing is needed anyway — the bundled
> copies cover the gap; if `check_latex()` still reports a missing package,
> install a user-space TinyTeX you control:
> `tinytex::install_tinytex(bundle = "TinyTeX")`, then restart R. On images
> whose TeX Live is frozen on a pre-2023 kernel (too old for `tabularray`),
> `emit()` skips LaTeX automatically and compiles through typst instead.

For decimal alignment in paper backends, metric-compatible fonts matter — check
with `check_fonts(spec)`.

## Troubleshooting

- **A relative DOCX path works.** `emit(spec, "out/x.docx")` resolves the path
  against your working directory like every other backend (the output path is
  absolutised before the OOXML zip is staged). No `normalizePath()` dance is
  needed.
- **If a PDF build appears to hang,** it is the LaTeX engine stopping at an
  interactive error prompt — fix the underlying LaTeX dependency (run
  `check_latex()`), or sidestep TeX entirely with
  `emit(spec, "out.pdf", format = "typst")`; render RTF/HTML to keep working
  in the meantime.
- **If a typst-compiled PDF renders in an unexpected font,** run
  `check_typst()` — it names the first available family of the chain, i.e.
  the face typst actually uses. Missing *later* members of the built-in
  fallback chain are normal cross-OS variance and are not warned about;
  only a family you explicitly named triggers the `emit()` warning.

## Cross-backend qualification (CDISC pilot)

The package ships a qualification (`inst/qualification/`) that rebuilds
representative CDISC-pilot tables (demographics, populations, AE overview, AE by
SOC/PT) from the public PHUSE Test Data Factory ADaM and renders **each to every
backend**, checking three things per cell: it emits without error, the file is
structurally valid, and an **independent count computed from the ADaM appears in
the rendered text of that backend** (cross-backend content parity). The four
pilot tables across all four file backends give a 16/16 PASS matrix:

| Table | RTF | HTML | DOCX | PDF¹ |
|---|:--:|:--:|:--:|:--:|
| 14-2.01 Demographics | PASS | PASS | PASS | PASS |
| 14-1.01 Populations | PASS | PASS | PASS | PASS |
| 14-3.01 TEAE overview | PASS | PASS | PASS | PASS |
| 14-3.04 TEAE by SOC/PT | PASS | PASS | PASS | PASS |

¹ The PDF column is verified **manually in a local environment** with a LaTeX
engine (`check_latex()`) and `pdftotext` for the text-parity check. It is **not
run in continuous integration**, which carries no TeX install; CI exercises the
RTF, HTML, and DOCX backends.

The runnable script (`inst/qualification/qualify_tabular_cdisc.R`) and how to
fetch the data are in the qualification README in that same `inst/qualification/`
folder; it is the most direct evidence that one `tabular` spec produces
consistent, correct output across all backends.
