---
title: "Output & qualification: backends, requirements, and the CDISC pilot"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

This article is about *rendering and proving* — choosing a backend, meeting its
system requirements, and the cross-backend validation. It does not cover building
or styling a table (see the other articles).

## `emit()` and `as_grid()`

`emit()` writes a file, dispatching on the extension:

```{r eval = FALSE}
emit(spec, "table.rtf")    # RTF
emit(spec, "table.html")   # HTML
emit(spec, "table.docx")   # Word
emit(spec, "table.pdf")    # PDF (via LaTeX)
emit(spec, "table.md")     # Markdown
```

`as_grid(spec)` resolves the fully-laid-out grid **without** writing a file —
useful for testing or programmatic inspection.

## Backend capability matrix

One spec renders to every backend, but the page-oriented features differ:

| Capability | RTF | HTML | DOCX | PDF/LaTeX | MD |
|---|:--:|:--:|:--:|:--:|:--:|
| Vertical pagination | ✓ | n/a¹ | ✓ | ✓ | n/a |
| Horizontal panels (`panels=`) | ✓ | n/a¹ | ✓ | ✓ | n/a |
| Per-page running header/footer | ✓ | – | ✓ | ✓ | – |
| `subgroup()` per-page BigN | ✓ | row² | ✓ | ✓ | row² |
| Continuation marker | panels only | – | – | ✓ | – |
| Decimal alignment (NBSP) | ✓ | ✓ | ✓ | ✓ | ✓ |
| System dependency | none | none | none | LaTeX | none |

¹ HTML/MD are one continuous document; the browser repeats `<thead>` on print.
² On HTML/MD the per-page N renders as a row under each subgroup banner instead
of in the repeating header.

## System requirements

**RTF, HTML, DOCX, Markdown need nothing beyond the R package.** Only PDF has a
system dependency — a LaTeX engine plus the packages the backend uses
(`tabularray`, `ninecolors`, `siunitx`, `tex-gyre`, …). Check and install:

```{r eval = FALSE}
check_latex()                                       # reports what's missing + the exact command
tinytex::tlmgr_install(c("tabularray", "ninecolors", "siunitx", "tex-gyre"))
```

> **OS-managed TeX Live (RHEL/dnf, Debian/apt):** `tlmgr` is locked and refuses
> to install ("will likely destroy the … TeXLive install"). Do **not** force it
> with `--ignore-warning`. Install a user-space TinyTeX you control instead:
> `tinytex::install_tinytex()`, restart R, then `tlmgr_install(...)`.

For decimal alignment in paper backends, metric-compatible fonts matter — check
with `check_fonts()`.

## Troubleshooting

- **A relative DOCX path works.** `emit(spec, "out/x.docx")` resolves the path
  against your working directory like every other backend (the output path is
  absolutised before the OOXML zip is staged). No `normalizePath()` dance is
  needed.
- **If a PDF build appears to hang,** it is the LaTeX engine stopping at an
  interactive error prompt — fix the underlying LaTeX dependency (run
  `check_latex()`); render RTF/HTML to keep working in the meantime.

## Cross-backend qualification (CDISC pilot)

The package ships a qualification (`inst/qualification/`) that rebuilds
representative CDISC-pilot tables (demographics, populations, AE overview, AE by
SOC/PT) from the public PHUSE Test Data Factory ADaM and renders **each to every
backend**, checking three things per cell: it emits without error, the file is
structurally valid, and an **independent count computed from the ADaM appears in
the rendered text of that backend** (cross-backend content parity). The four
pilot tables across RTF, HTML, and DOCX (the backends needing no system
dependency) give a 12/12 PASS matrix:

| Table | RTF | HTML | DOCX |
|---|:--:|:--:|:--:|
| 14-2.01 Demographics | PASS | PASS | PASS |
| 14-1.01 Populations | PASS | PASS | PASS |
| 14-3.01 TEAE overview | PASS | PASS | PASS |
| 14-3.04 TEAE by SOC/PT | PASS | PASS | PASS |

PDF renders identically once a LaTeX engine is available (`check_latex()`). The
runnable script (`qualify_tabular_cdisc.R`) and how to fetch the data are in the
qualification README; it is the most direct evidence that one `tabular` spec
produces consistent, correct output across all backends.
