---
title: "Presentation: titles, footnotes, page chrome, and styling"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
data(cdisc_saf_demo, package = "tabular")
data(cdisc_saf_n, package = "tabular")
N <- stats::setNames(cdisc_saf_n$n, cdisc_saf_n$arm_short)
arms <- c("placebo", "drug_50", "drug_100", "Total")
base <- tabular(cdisc_saf_demo) |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = "")
  ) |>
  cols_apply(arms, col_spec(align = "decimal"))
```

This article is about the *look* — titles, footnotes, the running page header /
footer, and cell styling. It assumes the table's shape is already built (see
[Structure](structure.html)) and never explains data prep or pagination for its
own sake.

## Titles and footnotes

Multi-line titles and static footnotes are arguments to `tabular()`:

```{r}
tabular(
  cdisc_saf_demo,
  titles = c(
    "Table 14-2.01",
    "Demographic and Baseline Characteristics",
    "ITT Population"
  ),
  footnotes = c(
    "Percentages are based on the number of ITT subjects per arm.",
    "BMI = body mass index."
  )
) |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = "")
  ) |>
  cols_apply(arms, col_spec(align = "decimal"))
```

For an **anchored** footnote (an auto-numbered superscript on a specific cell or
header) use `footnote()` with a `cells_*()` location:

```{r}
base |>
  footnote(
    "Excludes one subject withdrawn before dosing.",
    .at = cells_headers(j = "Total")
  )
```

## The running page header / footer

Regulatory TFLs carry a header on every page (protocol, page X of Y, data-cut).
That is preset page chrome, not a title. `pagehead` / `pagefoot` each take
`left` / `right` (and `center`) vectors; the tokens `{page}`, `{npages}`,
`{program}`, `{datetime}` resolve at render time:

```{r}
chrome <- base |>
  preset(
    pagehead = list(
      left = c("Analysis Set: Safety", "Protocol: XYZ-123"),
      right = c("Data cut: 2026-01-15", "Page {page} of {npages}")
    )
  )
chrome
```

The running header/footer is page chrome — it repeats on every page of the
**paged backends** (RTF, PDF, DOCX) and does not appear in the single-page HTML
preview above. Emit to a paged backend to see it:

```{r eval = FALSE}
emit(chrome, "demographics.pdf") # protocol + page x of y on every page
```

> **Stacking direction.** Each vector stacks **outward from the table**: index 1
> is the line nearest the table body, later elements move toward the page edge.
> For a header that reads "Analysis Set" on top and "Protocol" just above the
> table, put `"Protocol…"` first. (Easy to invert — check the rendered page.)

## Cell styling

`style()` appends one location + attribute layer. Target a region with a
`cells_*()` constructor; attributes include `bold`, `italic`, `underline`,
`color`, `background`, and borders via `brdr()`:

```{r}
base |>
  # the column-header band is bold by default, so style it with colour /
  # background for a visible change rather than re-applying bold
  style(color = "#1F3B5C", background = "#DBE4F0", .at = cells_headers()) |>
  style(italic = TRUE, .at = cells_group_headers()) |> # italic section rows
  style(background = "#F2F2F2", .at = cells_body(j = "Total")) # shade the Total column
```

Body filters live on `cells_body()`: `i =` (row index), `j =` (column name),
`where =` (an expression over the data).

A border attribute takes a `brdr()` value — width, line style, colour. Here a
hairline rule separates the pooled Total column from the per-arm columns:

```{r}
base |>
  style(
    border_left = brdr("hairline", "solid", "#ADB5BD"),
    .at = cells_body(j = "Total")
  )
```

`border` styles all four sides at once; `border_top` / `border_bottom` /
`border_left` / `border_right` target one. The table-wide rules (toprule,
midrule, bottomrule) are preset territory — see the `rules` knob on
`preset()`.

## Inline markup

Plain strings render as plain text — a stray `**` never silently bolds the
surrounding span. Wrap a string in `md()` (CommonMark plus `^sup^` /
`~sub~`) or `html()` (a constrained inline tag set) to opt in, the same
convention gt uses. Every string slot accepts them: titles, footnotes,
`col_spec()` labels, and `style()` pretext / posttext:

```{r}
tabular(
  cdisc_saf_demo,
  titles = c(
    "Table 14-2.01",
    "Demographic and Baseline Characteristics",
    md("*ITT Population*")
  ),
  footnotes = c(
    md("BMI = body mass index (kg/m^2^)."),
    html("Source: <i>ADSL</i>, data cut 15JAN2026.")
  )
) |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = "")
  ) |>
  cols_apply(arms, col_spec(align = "decimal"))
```

The marked string survives `c()` concatenation, so plain and marked lines mix
freely in one titles or footnotes vector. Raw HTML inside `md()` passes
through the same tag whitelist as `html()`; tags outside it drop their
wrapper and keep the text.

## Presets: cosmetics and fit

`preset()` carries the cosmetic defaults — fonts, rules, padding, `na_text`,
paper/orientation/margins — and the **`width_mode`** that decides how the table
fills the page:

| `width_mode` | Effect |
|---|---|
| `"content"` *(default)* | columns sized to their content |
| `"window"` | `"auto"` columns **stretch to fill** the printable width (Word "Auto-fit Window") |
| `"fixed"` | only the widths you pin are used |

```{r}
base |>
  preset(
    font_size = 9,
    orientation = "landscape",
    paper_size = "letter",
    margins = c(1, 0.75, 1, 0.75),
    width_mode = "window",
    na_text = "-"
  )
```

Use `set_preset()` once at the top of a study to make these defaults apply to
every table without restating them.

> **Decimal alignment** (`col_spec(align = "decimal")`) pads numeric cells with
> non-breaking spaces so the decimal points line up. Alignment is exact in a
> monospaced font; in proportional fonts it is visually close (em-aware padding
> is future work). For submission TFLs in Courier this is exact.
