---
title: "Presentation: titles, footnotes, page chrome, and styling"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
data(cdisc_saf_demo, package = "tabular")
data(cdisc_saf_n, package = "tabular")
N <- stats::setNames(cdisc_saf_n$n, cdisc_saf_n$arm_short)
arms <- c("placebo", "drug_50", "drug_100", "Total")
base <- tabular(cdisc_saf_demo) |>
  cols(variable = col_spec(usage = "group", group_display = "header_row", label = ""),
       stat_label = col_spec(usage = "indent", label = "")) |>
  cols_apply(arms, col_spec(align = "decimal"))
```

This article is about the *look* — titles, footnotes, the running page header /
footer, and cell styling. It assumes the table's shape is already built (see
[Structure](structure.html)) and never explains data prep or pagination for its
own sake.

## Titles and footnotes

Multi-line titles and static footnotes are arguments to `tabular()`:

```{r}
tabular(
  cdisc_saf_demo,
  titles    = c("Table 14-2.01",
                "Demographic and Baseline Characteristics", "ITT Population"),
  footnotes = c("Percentages are based on the number of ITT subjects per arm.",
                "BMI = body mass index.")
) |>
  cols(variable = col_spec(usage = "group", group_display = "header_row", label = ""),
       stat_label = col_spec(usage = "indent", label = "")) |>
  cols_apply(arms, col_spec(align = "decimal")) |>
  emit(tempfile(fileext = ".html")) |> invisible()
```

For an **anchored** footnote (an auto-numbered superscript on a specific cell or
header) use `footnote()` with a `cells_*()` location:

```{r}
base |>
  footnote("Excludes one subject withdrawn before dosing.",
           .at = cells_headers(j = "Total")) |>
  emit(tempfile(fileext = ".html")) |> invisible()
```

## The running page header / footer

Regulatory TFLs carry a header on every page (protocol, page X of Y, data-cut).
That is preset page chrome, not a title. `pagehead` / `pagefoot` each take
`left` / `right` (and `center`) vectors; the tokens `{page}`, `{npages}`,
`{program}`, `{datetime}` resolve at render time:

```{r}
base |>
  preset(
    pagehead = list(
      left  = c("Analysis Set: Safety", "Protocol: XYZ-123"),
      right = c("Data cut: 2026-01-15", "Page {page} of {npages}")
    )
  ) |>
  emit(tempfile(fileext = ".html")) |> invisible()
```

> **Stacking direction.** Each vector stacks **outward from the table**: index 1
> is the line nearest the table body, later elements move toward the page edge.
> For a header that reads "Analysis Set" on top and "Protocol" just above the
> table, put `"Protocol…"` first. (Easy to invert — check the rendered page.)

## Cell styling

`style()` appends one location + attribute layer. Target a region with a
`cells_*()` constructor; attributes include `bold`, `italic`, `underline`,
`color`, `background`, and borders via `brdr()`:

```{r}
base |>
  style(bold = TRUE, .at = cells_headers()) |>                 # bold the header band
  style(italic = TRUE, .at = cells_group_headers()) |>         # italic section rows
  style(background = "#F2F2F2", .at = cells_body(j = "Total")) |>  # shade the Total column
  emit(tempfile(fileext = ".html")) |> invisible()
```

Body filters live on `cells_body()`: `i =` (row index), `j =` (column name),
`where =` (an expression over the data).

## Presets: cosmetics and fit

`preset()` carries the cosmetic defaults — fonts, rules, padding, `na_text`,
paper/orientation/margins — and the **`width_mode`** that decides how the table
fills the page:

| `width_mode` | Effect |
|---|---|
| `"content"` *(default)* | columns sized to their content |
| `"window"` | `"auto"` columns **stretch to fill** the printable width (Word "Auto-fit Window") |
| `"fixed"` | only the widths you pin are used |

```{r}
base |>
  preset(font_size = 9, orientation = "landscape", paper_size = "letter",
         margins = c(1, 0.75, 1, 0.75), width_mode = "window", na_text = "-") |>
  emit(tempfile(fileext = ".html")) |> invisible()
```

Use `set_preset()` once at the top of a study to make these defaults apply to
every table without restating them.

> **Decimal alignment** (`col_spec(align = "decimal")`) pads numeric cells with
> non-breaking spaces so the decimal points line up. Alignment is exact in a
> monospaced font; in proportional fonts it is visually close (em-aware padding
> is future work). For submission TFLs in Courier this is exact.
