---
title: "Recipes: the CDISC-pilot tables, end to end"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

These are the canonical safety and efficacy tables a programmer builds from the
shell — each one rendered **live** below from the bundled demo data, so what you
see is exactly what the code produces. They mirror the package's cross-backend
qualification (`inst/qualification/`), which rebuilds the same displays from the
real PHUSE Test Data Factory ADaM and checks them across every backend.

Each recipe ends on the spec (the live HTML you see); the closing line shows how
to **ship the same spec** to a submission backend — just change the extension.

## Demographics and baseline characteristics

The reference safety table: a `usage = "group"` parameter stub, indented
statistic rows, decimal-aligned arm columns, and the population BigN folded into
each header from `cdisc_saf_n`.

```{r}
n <- stats::setNames(cdisc_saf_n$n, cdisc_saf_n$arm_short)

demo_tbl <- tabular(
  cdisc_saf_demo,
  titles = c(
    "Table 14.1.1",
    "Demographic and Baseline Characteristics",
    "Safety Population"
  ),
  footnotes = "Percentages are based on the number of subjects per treatment group."
) |>
  cols(
    variable = col_spec(usage = "group", label = "Parameter"),
    stat_label = col_spec(label = "Statistic"),
    placebo = col_spec(
      label = "Placebo\nN={n['placebo']}",
      align = "decimal"
    ),
    drug_50 = col_spec(
      label = "Drug 50\nN={n['drug_50']}",
      align = "decimal"
    ),
    drug_100 = col_spec(
      label = "Drug 100\nN={n['drug_100']}",
      align = "decimal"
    ),
    Total = col_spec(label = "Total\nN={n['Total']}", align = "decimal")
  )

demo_tbl
```

```{r eval = FALSE}
emit(demo_tbl, "t_14_1_1.rtf") # ship: or .pdf / .docx / .html / .md
```

## Adverse-event overview

A flat overview: `stat_label` is a plain display column carrying both the
high-level flag rows and the two-space-indented maximum-severity detail rows
(the indent is baked into the label strings), so each category sits on one line
with its counts.

```{r}
ae_tbl <- tabular(
  cdisc_saf_ae,
  titles = c(
    "Table 14.3.0",
    "Overview of Treatment-Emergent Adverse Events",
    "Safety Population"
  )
) |>
  cols(
    stat_label = col_spec(label = ""),
    placebo = col_spec(
      label = "Placebo\nN={n['placebo']}",
      align = "decimal"
    ),
    drug_50 = col_spec(
      label = "Drug 50\nN={n['drug_50']}",
      align = "decimal"
    ),
    drug_100 = col_spec(
      label = "Drug 100\nN={n['drug_100']}",
      align = "decimal"
    ),
    Total = col_spec(label = "Total\nN={n['Total']}", align = "decimal")
  )

ae_tbl
```

```{r eval = FALSE}
emit(ae_tbl, "t_14_3_0.rtf") # ship: or .pdf / .docx / .html / .md
```

## Adverse events by SOC and preferred term

The two-level hierarchy: `label` holds SOC text on SOC rows and PT text on PT
rows, indented by `indent_level`; the hidden `soc` / `row_type` / `n_total` /
`soc_n` columns ride along as partition and sort keys. `sort_rows()` clusters
PTs under their parent SOC and orders both levels by descending frequency.

```{r}
aesocpt_tbl <- tabular(
  cdisc_saf_aesocpt,
  titles = c(
    "Table 14.3.1",
    "Adverse Events by System Organ Class and Preferred Term",
    "Safety Population"
  )
) |>
  cols(
    label = col_spec(
      label = "SOC / Preferred Term",
      indent_by = "indent_level"
    ),
    soc = col_spec(visible = FALSE),
    row_type = col_spec(visible = FALSE),
    n_total = col_spec(visible = FALSE),
    soc_n = col_spec(visible = FALSE),
    placebo = col_spec(
      label = "Placebo\nN={n['placebo']}",
      align = "decimal"
    ),
    drug_50 = col_spec(
      label = "Drug 50\nN={n['drug_50']}",
      align = "decimal"
    ),
    drug_100 = col_spec(
      label = "Drug 100\nN={n['drug_100']}",
      align = "decimal"
    ),
    Total = col_spec(label = "Total\nN={n['Total']}", align = "decimal")
  ) |>
  sort_rows(by = c("soc_n", "n_total"), descending = c(TRUE, TRUE))

aesocpt_tbl
```

This table is long, so a real deliverable paginates it — keeping each SOC and its
preferred terms together. Pagination materialises in the paged backends (RTF,
PDF, DOCX), so add `paginate()` and emit to one of those:

```{r eval = FALSE}
aesocpt_tbl |>
  paginate(keep_together = "soc", continuation = "(continued)") |>
  emit("t_14_3_1.pdf")
```

## Best overall response and response rates

The efficacy companion: `group_label` synthesises one bold section band per
analysis block (best overall response, ORR, CBR, DCR) via `header_row`, with
indented response rows beneath. Denominators come from `cdisc_eff_n`.

```{r}
ne <- stats::setNames(cdisc_eff_n$n, cdisc_eff_n$arm_short)

resp_tbl <- tabular(
  cdisc_eff_resp,
  titles = c(
    "Table 14.2.1",
    "Best Overall Response and Response Rates",
    "Efficacy Evaluable Population"
  )
) |>
  cols(
    group_label = col_spec(usage = "group", group_display = "header_row"),
    stat_label = col_spec(label = "Response"),
    groupid = col_spec(visible = FALSE),
    row_type = col_spec(visible = FALSE),
    placebo = col_spec(
      label = "Placebo\nN={ne['placebo']}",
      align = "decimal"
    ),
    drug_50 = col_spec(
      label = "Drug 50\nN={ne['drug_50']}",
      align = "decimal"
    ),
    drug_100 = col_spec(
      label = "Drug 100\nN={ne['drug_100']}",
      align = "decimal"
    )
  )

resp_tbl
```

```{r eval = FALSE}
emit(resp_tbl, "t_14_2_1.rtf") # ship: or .pdf / .docx / .html / .md
```

## See also

- [Data in](data-in.html) — produce these wide frames from a cards/cardx ARD with
  `pivot_across()`.
- [Structure](structure.html) and [Presentation](presentation.html) — the column,
  pagination, and styling verbs each recipe draws on.
- [Output & qualification](output.html) — the cross-backend validation these
  recipes mirror.
