---
title: "Structure: columns, headers, and pagination"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

This article is about *shape*: which column does what, multi-level headers, and
how a table that is too long or too wide is split across pages. It assumes you
already have a wide frame (see [Data in](data-in.html)) and does not cover
cosmetics (see [Presentation](presentation.html)).

## Row grouping: `group_rows()`

Row structure is a fact about the whole table, so it is declared once with
`group_rows()` — not per column. `by` names the grouping key columns, ordered
outer to inner; `display` picks how each level renders:

| `display` | Use it for | Behaviour |
|---|---|---|
| `"header_row"` *(default)* | section variable (e.g. parameter) | each value becomes a **section-header row**; the key column is hidden |
| `"column"` | a visible row label | column stays; repeated values are suppressed |
| `"column_repeat"` | a visible row label | column stays; every row repeats the value |
| `"none"` | a hidden block key | break-only: no header, no column — just group transitions |

`cols()` handles the per-column cosmetics — labels (`x = "Label"` is shorthand
for `col_spec(label = )`), alignment, widths, and `.hide =` for helper columns.
Indentation is the separate `col_spec(indent = …)` argument (a fixed integer
level, or a column name for per-row depth).

```{r}
data(cdisc_saf_demo, package = "tabular")
arms <- c("placebo", "drug_50", "drug_100", "Total")

tabular(cdisc_saf_demo, titles = "Demographics") |>
  cols(variable = "", stat_label = "") |>
  cols_apply(arms, col_spec(align = "decimal")) |>
  group_rows(by = "variable")
```

`cols_apply()` attaches one shared `col_spec` to **all** the arm columns at once —
use it instead of repeating `cols(placebo = …, drug_50 = …)` for a variable
number of arms.

> **Indent from exactly one source.** `display = "header_row"` already
> indents its child rows one level, so the stub column (here `stat_label`) needs
> **no** `indent` — the section supplies it. (An explicit `indent` on the host
> *overrides* that auto-indent rather than stacking, so `indent = 1` there still
> yields a single level.) The same care applies to labels from `pivot_across()`,
> which come out with a leading indent baked into the string: keep them as-is or
> `trimws()` them and set `indent` yourself — don't double up.

## BigN in the column headers

The `(N=…)` denominator goes in each arm's header label. Build it from a BigN
table and interpolate with glue:

```{r}
data(cdisc_saf_n, package = "tabular")
N <- stats::setNames(cdisc_saf_n$n, cdisc_saf_n$arm_short)

tabular(cdisc_saf_demo, titles = "Demographics") |>
  group_rows(by = "variable") |>
  cols(
    stat_label = col_spec(label = ""),
    placebo = col_spec(
      label = "Placebo\n(N={N['placebo']})",
      align = "decimal"
    ),
    drug_50 = col_spec(
      label = "Drug 50\n(N={N['drug_50']})",
      align = "decimal"
    ),
    drug_100 = col_spec(
      label = "Drug 100\n(N={N['drug_100']})",
      align = "decimal"
    ),
    Total = col_spec(label = "Total\n(N={N['Total']})", align = "decimal")
  )
```

> **Clinical convention:** BigN is the population denominator (from ADSL), **not**
> the number of rows in the domain dataset — compute it from the population, not
> from the summarised data.

For a **variable** number of arms, the per-arm label is one `cols_apply()` call
instead of a hand-written line each: the `{.name}` token resolves to each matched
column's name, and the rest of the `{…}` evaluates in the calling environment, so
the BigN looks itself up:

```{r}
arm_cols <- c("placebo", "drug_50", "drug_100", "Total")

tabular(cdisc_saf_demo, titles = "Demographics") |>
  group_rows(by = "variable") |>
  cols(stat_label = col_spec(label = "")) |>
  cols_apply(
    arm_cols,
    col_spec(label = "{.name}\n(N={N[.name]})", align = "decimal")
  )
```

## Multi-level headers and widths

`headers()` builds spanning bands over groups of columns:

```{r}
tabular(cdisc_saf_demo, titles = "Demographics") |>
  group_rows(by = "variable") |>
  cols(stat_label = col_spec(label = "", width = "2.2in")) |>
  cols_apply(arms, col_spec(align = "decimal", width = "1in")) |>
  headers("Treatment Group" = c("placebo", "drug_50", "drug_100", "Total"))
```

Widths: `"auto"` (default) sizes to content; a pinned value (`"1in"`, `1.0`,
`"20%"`) wraps within that width. **Set the shared arm width via `cols_apply()`
last** — its non-default `width` then wins the field-merge; a later `cols()` call
carrying the default `width = "auto"` would otherwise be ambiguous.

## Sorting rows

Display cells are formatted strings — `"54 (21.3)"` sorts lexically, not
numerically. The idiom: carry one hidden **numeric key** per sort level, hide
it with `col_spec(visible = FALSE)`, and hand the keys to `sort_rows()`.
`descending` takes one value per key, so mixed-direction sorts are a single
call.

The bundled AE table ships its keys precomputed: `soc_n` (events in the parent
SOC, constant down each SOC block) and `n_total` (events on the row). Sorting
on both, descending, clusters every preferred term under its SOC and orders
both levels by frequency — the standard SAP ordering:

```{r}
data(cdisc_saf_aesocpt, package = "tabular")

tabular(cdisc_saf_aesocpt, titles = "AEs by SOC and PT, descending frequency") |>
  cols(
    label = col_spec(
      label = "SOC / Preferred Term",
      indent = "indent_level"
    ),
    .hide = c("soc", "row_type", "n_total", "soc_n")
  ) |>
  cols_apply(arms, col_spec(align = "decimal")) |>
  sort_rows(by = c("soc_n", "n_total"), descending = c(TRUE, TRUE))
```

Because `soc_n` is constant within a SOC block and never smaller than any PT's
`n_total` inside it, each SOC's summary row sorts to the top of its own block —
no separate "parent first" switch needed.

## Pagination — long tables

`paginate()` derives the rows-per-page budget from the preset (paper, font,
margins) and the title/footnote/header line counts — you never set rows-per-page
by hand. `keep_together` stops a page break landing inside a section's run:

```{r}
data(cdisc_saf_aesocpt, package = "tabular")
ae_pages <- tabular(cdisc_saf_aesocpt, titles = "AEs by SOC and PT") |>
  cols(
    label = col_spec(
      label = "SOC / Preferred Term",
      indent = "indent_level"
    ),
    .hide = c("soc", "row_type", "n_total", "soc_n")
  ) |>
  cols_apply(
    c("placebo", "drug_50", "drug_100", "Total"),
    col_spec(align = "decimal")
  ) |>
  paginate(
    keep_together = "soc",
    orphan_floor = 4,
    widow_floor = 2,
    continuation = "(continued)"
  )
ae_pages
```

The preview above is one continuous table: row pagination, `keep_together`, and
the `continuation` marker materialise only in the **paged backends** (RTF, PDF,
DOCX), not in HTML. Emit to one of those to see the page breaks:

```{r eval = FALSE}
emit(ae_pages, "ae_soc_pt.pdf") # continuation marker repeats on each continued page
```

## Panels — wide tables

When the columns don't fit one page, `paginate(panels = N)` splits the
**non-stub** columns into `N` chunks and repeats the stub on each panel (so the
row labels reappear). The stub defaults to the `group_rows()` keys; name it
explicitly with `repeat_cols` to carry the row label too:

```{r}
wide_split <- tabular(cdisc_saf_demo, titles = "Demographics (wide split)") |>
  cols(variable = "", stat_label = "") |>
  cols_apply(arms, col_spec(align = "decimal")) |>
  group_rows(by = "variable") |>
  paginate(
    panels = 2,
    repeat_cols = c("variable", "stat_label"), # both ride every panel
    continuation = "(continued)"
  )
wide_split
```

Panels are a paged-backend feature: in HTML and Markdown the table stays one
continuous block (the preview above), while RTF, PDF, and DOCX place each panel
on its own page with the stub columns repeated. Emit to a paged backend to see
the split:

```{r eval = FALSE}
emit(wide_split, "demographics_wide.pdf") # panel 2 carries the (continued) marker
```

Two things to know:

- **`panels = N` splits into `N` *equal* chunks** — there is no explicit split
  position (no "first 5, then the rest"). Equal split is fine for page-fit; if
  you need a specific boundary, that is a known limitation.
- **`panels` is a positive integer** (default `1` = no split). Width-aware
  automatic splitting is a planned future feature, not a current option.

## Subgroups and per-page BigN

`subgroup()` partitions the table — one page block per value, with a banner and a
hard page break. A partition-constant column can ride into the banner:

```{r}
data(cdisc_saf_subgroup, package = "tabular")
tabular(cdisc_saf_subgroup, titles = "Vital signs by sex") |>
  cols(
    param = "Parameter",
    visit = "Visit",
    stat_label = "Statistic",
    .hide = c("sex", "sex_n", "paramcd")
  ) |>
  group_rows(by = c("param", "visit")) |>
  cols_apply(
    c("placebo", "drug_50", "drug_100", "Total"),
    col_spec(align = "decimal")
  ) |>
  subgroup(by = "sex", label = "Sex: {sex} (N = {sex_n})") # page total in banner
```

For a **different `(N=)` per arm on each page** (the column headers re-resolving
per subgroup), pass `big_n` — a small table of N per page × arm. No bundled
dataset carries per-arm-per-page counts, so build it inline (this is also the
shape `big_n` expects):

```{r}
big_n <- tibble::tribble(
  ~sex, ~placebo, ~drug_50, ~drug_100, ~Total,
  "F",       53L,      55L,       35L,    143L,
  "M",       33L,      41L,       37L,    111L
)

tabular(cdisc_saf_subgroup, titles = "Vital signs by sex") |>
  cols(
    param = "Parameter",
    visit = "Visit",
    stat_label = "Statistic",
    .hide = c("sex_n", "paramcd")
  ) |>
  group_rows(by = c("param", "visit")) |>
  cols_apply(
    c("placebo", "drug_50", "drug_100", "Total"),
    col_spec(align = "decimal")
  ) |>
  subgroup(by = "sex", label = "Sex: {sex}", big_n = big_n) # per-page (N=) per arm
```

`big_n` accepts this wide shape (page column + one column per arm) or a long
`count()`-style table (page, arm, n).

## Empty tables: no data to report

A table whose data has zero rows still renders — the full page chrome and the
column headers stay intact, and an empty-data placeholder takes the place of the
body. This is the correct output for a population that produced no records (a
cohort with no subjects, a serious-AE table with no events), rather than an error
or a blank page. The per-table message is `tabular(empty_text = ...)`; set a
house default for every table with `preset(empty_text = ...)`. The message
renders as a single horizontally centred row in the table body, where the first
data row would sit.

```{r}
# Same demographics shell, but the population filter has left no rows.
empty_demo <- cdisc_saf_demo[0, ]

tabular(
  empty_demo,
  titles = c(
    "Table 14.1.1",
    "Demographic and Baseline Characteristics",
    "Safety Population"
  ),
  footnotes = "No subjects met the inclusion criteria for this cohort.",
  empty_text = "No data available to report"
) |>
  group_rows(by = "variable") |>
  cols(
    stat_label = col_spec(label = "Statistic"),
    placebo = col_spec(label = "Placebo", align = "decimal"),
    drug_50 = col_spec(label = "Drug 50 mg", align = "decimal"),
    drug_100 = col_spec(label = "Drug 100 mg", align = "decimal"),
    Total = col_spec(align = "decimal")
  )
```

The same applies under `subgroup()`: a zero-N crossing is dropped by default, but
`subgroup(..., keep_empty = TRUE)` keeps it and renders its banner above the
empty-data page — so every level in the shell appears even when one has no data.
