---
title: "Structure: columns, headers, and pagination"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tabular)
```

This article is about *shape*: which column does what, multi-level headers, and
how a table that is too long or too wide is split across pages. It assumes you
already have a wide frame (see [Data in](data-in.html)) and does not cover
cosmetics (see [Presentation](presentation.html)).

## The column model: `usage`

Every column gets a role via `col_spec(usage = …)`. Picking the right one is the
single most important structural decision:

| `usage` | Use it for | Behaviour |
|---|---|---|
| `"display"` *(default)* | data cells (the arm columns) | one value per row |
| `"group"` | section variable (e.g. parameter) | each value becomes a **section-header row**; the column is hidden |
| `"indent"` | manual extra nesting (rarely needed) | prefixes one indent level. **`group_display = "header_row"` already indents its child rows**, so don't add this on top |
| `"id"` | the row label that must stay visible | like `display`, but **joins the stub and repeats on every horizontal panel** |

```{r}
data(cdisc_saf_demo, package = "tabular")
arms <- c("placebo", "drug_50", "drug_100", "Total")

tabular(cdisc_saf_demo, titles = "Demographics") |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = "")
  ) |>
  cols_apply(arms, col_spec(align = "decimal"))
```

`cols_apply()` attaches one shared `col_spec` to **all** the arm columns at once —
use it instead of repeating `cols(placebo = …, drug_50 = …)` for a variable
number of arms.

> **Indent from exactly one source.** `group_display = "header_row"` already
> indents its child rows one level, so the stub column (here `stat_label`) needs
> **no** `indent` — the section supplies it. (An explicit `indent` on the host
> *overrides* that auto-indent rather than stacking, so `indent = 1` there still
> yields a single level.) The same care applies to labels from `pivot_across()`,
> which come out with a leading indent baked into the string: keep them as-is or
> `trimws()` them and set `indent` yourself — don't double up.

## BigN in the column headers

The `(N=…)` denominator goes in each arm's header label. Build it from a BigN
table and interpolate with glue:

```{r}
data(cdisc_saf_n, package = "tabular")
N <- stats::setNames(cdisc_saf_n$n, cdisc_saf_n$arm_short)

tabular(cdisc_saf_demo, titles = "Demographics") |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = ""),
    placebo = col_spec(
      label = "Placebo\n(N={N['placebo']})",
      align = "decimal"
    ),
    drug_50 = col_spec(
      label = "Drug 50\n(N={N['drug_50']})",
      align = "decimal"
    ),
    drug_100 = col_spec(
      label = "Drug 100\n(N={N['drug_100']})",
      align = "decimal"
    ),
    Total = col_spec(label = "Total\n(N={N['Total']})", align = "decimal")
  )
```

> **Clinical convention:** BigN is the population denominator (from ADSL), **not**
> the number of rows in the domain dataset — compute it from the population, not
> from the summarised data.

For a **variable** number of arms, the per-arm label is one `cols_apply()` call
instead of a hand-written line each: the `{.name}` token resolves to each matched
column's name, and the rest of the `{…}` evaluates in the calling environment, so
the BigN looks itself up:

```{r}
arm_cols <- c("placebo", "drug_50", "drug_100", "Total")

tabular(cdisc_saf_demo, titles = "Demographics") |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = "")
  ) |>
  cols_apply(
    arm_cols,
    col_spec(label = "{.name}\n(N={N[.name]})", align = "decimal")
  )
```

## Multi-level headers and widths

`headers()` builds spanning bands over groups of columns:

```{r}
tabular(cdisc_saf_demo, titles = "Demographics") |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(label = "", width = "2.2in")
  ) |>
  cols_apply(arms, col_spec(align = "decimal", width = "1in")) |>
  headers("Treatment Group" = c("placebo", "drug_50", "drug_100", "Total"))
```

Widths: `"auto"` (default) sizes to content; a pinned value (`"1in"`, `1.0`,
`"20%"`) wraps within that width. **Set the shared arm width via `cols_apply()`
last** — its non-default `width` then wins the field-merge; a later `cols()` call
carrying the default `width = "auto"` would otherwise be ambiguous.

## Pagination — long tables

`paginate()` derives the rows-per-page budget from the preset (paper, font,
margins) and the title/footnote/header line counts — you never set rows-per-page
by hand. `keep_together` stops a page break landing inside a section's run:

```{r}
data(cdisc_saf_aesocpt, package = "tabular")
ae_pages <- tabular(cdisc_saf_aesocpt, titles = "AEs by SOC and PT") |>
  cols(
    label = col_spec(
      label = "SOC / Preferred Term",
      indent = "indent_level"
    ),
    soc = col_spec(
      usage = "group",
      visible = FALSE,
      group_display = "column_repeat"
    ),
    row_type = col_spec(visible = FALSE),
    n_total = col_spec(visible = FALSE),
    soc_n = col_spec(visible = FALSE)
  ) |>
  cols_apply(
    c("placebo", "drug_50", "drug_100", "Total"),
    col_spec(align = "decimal")
  ) |>
  paginate(
    keep_together = "soc",
    orphan_floor = 4,
    widow_floor = 2,
    continuation = "(continued)"
  )
ae_pages
```

The preview above is one continuous table: row pagination, `keep_together`, and
the `continuation` marker materialise only in the **paged backends** (RTF, PDF,
DOCX), not in HTML. Emit to one of those to see the page breaks:

```{r eval = FALSE}
emit(ae_pages, "ae_soc_pt.pdf") # continuation marker repeats on each continued page
```

## Panels — wide tables

When the columns don't fit one page, `paginate(panels = N)` splits the
**non-group** columns into `N` chunks and repeats every `group`/`id` column on
each panel (so the row labels reappear). Make the row label `usage = "id"` so it
rides every panel:

```{r}
wide_split <- tabular(cdisc_saf_demo, titles = "Demographics (wide split)") |>
  cols(
    variable = col_spec(
      usage = "group",
      group_display = "header_row",
      label = ""
    ),
    stat_label = col_spec(usage = "id", label = "") # repeats on every panel
  ) |>
  cols_apply(arms, col_spec(align = "decimal")) |>
  paginate(panels = 2, continuation = "(continued)")
wide_split
```

Panels are a paged-backend feature: in HTML and Markdown the table stays one
continuous block (the preview above), while RTF, PDF, and DOCX place each panel
on its own page with the `id` / `group` columns repeated. Emit to a paged
backend to see the split:

```{r eval = FALSE}
emit(wide_split, "demographics_wide.pdf") # panel 2 carries the (continued) marker
```

Two things to know:

- **`panels = N` splits into `N` *equal* chunks** — there is no explicit split
  position (no "first 5, then the rest"). Equal split is fine for page-fit; if
  you need a specific boundary, that is a known limitation.
- **`panels = "auto"` is currently a no-op** (treated as `1`): it does *not*
  auto-fit, it just leaves every column on one page (and the table may overflow
  with a warning). Always pass an explicit integer.

## Subgroups and per-page BigN

`subgroup()` partitions the table — one page block per value, with a banner and a
hard page break. A partition-constant column can ride into the banner:

```{r}
data(cdisc_saf_subgroup, package = "tabular")
tabular(cdisc_saf_subgroup, titles = "Vital signs by sex") |>
  cols(
    sex = col_spec(visible = FALSE),
    sex_n = col_spec(visible = FALSE),
    paramcd = col_spec(visible = FALSE),
    param = col_spec(usage = "group", label = "Parameter"),
    visit = col_spec(usage = "group", label = "Visit"),
    stat_label = col_spec(usage = "id", label = "Statistic")
  ) |>
  cols_apply(
    c("placebo", "drug_50", "drug_100", "Total"),
    col_spec(align = "decimal")
  ) |>
  subgroup(by = "sex", label = "Sex: {sex} (N = {sex_n})") # page total in banner
```

For a **different `(N=)` per arm on each page** (the column headers re-resolving
per subgroup), pass `big_n` — a small table of N per page × arm. No bundled
dataset carries per-arm-per-page counts, so build it inline (this is also the
shape `big_n` expects):

```{r}
big_n <- tibble::tribble(
  ~sex, ~placebo, ~drug_50, ~drug_100, ~Total,
  "F",       53L,      55L,       35L,    143L,
  "M",       33L,      41L,       37L,    111L
)

tabular(cdisc_saf_subgroup, titles = "Vital signs by sex") |>
  cols(
    sex_n = col_spec(visible = FALSE),
    paramcd = col_spec(visible = FALSE),
    param = col_spec(usage = "group", label = "Parameter"),
    visit = col_spec(usage = "group", label = "Visit"),
    stat_label = col_spec(usage = "id", label = "Statistic")
  ) |>
  cols_apply(
    c("placebo", "drug_50", "drug_100", "Total"),
    col_spec(align = "decimal")
  ) |>
  subgroup(by = "sex", label = "Sex: {sex}", big_n = big_n) # per-page (N=) per arm
```

`big_n` accepts this wide shape (page column + one column per arm) or a long
`count()`-style table (page, arm, n).
